---
title: "OysterRNA24 DESeq2 Analysis"
author: "Callie Hundley"
format: 
  html:
    df-print: paged 
    code-fold: true 
    code-tools: true
editor: visual
---

# Project Overview

## Loading packages

```{r, warning=FALSE}
library(tidyverse)
library(lubridate)
library(readxl)
library(readr)
library(tibble)
# BiocManager::install("DESeq2") # <- install only once!
#library(DESeq2)
# BiocManager::install("pheatmap") # <- install only once 
#library(pheatmap)
# BiocManager::install("ComplexHeatmap")
library(ComplexHeatmap)
library(RColorBrewer)
library(ggplot2)
```

## Loading data

Start by loading in your data:

-   gene counts matrix generated by `featureCounts` at the end of the RNASeq pipeline --\> saved as `counts_data`
-   sample metadata --\> saved as `col_data`
-   gene annotations (information about each gene included as the rows in the `counts_data` dataframe) file from the GTF file associated with the genome assembly --\> saved as `gtf` (commonly saved as `rowranges` in other pipelines)

**Gene count data**

```{r}
# Read in the gene count data -- matrix format 
counts_data <- read_tsv("/work/clh162/OysterRNA24/hisat2_align/featureCounts_counts/gene_counts_matrix.txt") %>% 
  column_to_rownames(var = "Geneid") #assigning the gene names as the row names 
head(counts_data)

# 7/2/26 TEST -- Sophie's count data
counts_data <- read_tsv("/work/clh162/Data+/Sophie/2025-2026-team-code-sac165/Data+_team/Sophie/gene_counts_clean.txt") %>% 
  dplyr::rename(Geneid = 1) %>% 
  column_to_rownames(var = "Geneid") #assigning the gene names as the row names 
```

**Sample metadata**

```{r}
# Read in the sample data (defining sequenced samples)
coldata <- read_excel("/work/clh162/OysterRNA24/xRNA for BC Summer 2024.xlsx",
                      col_types = c("text", "text", "text", "text", "text", "date", "date", "numeric", 
                                    "numeric", "numeric", "numeric"),
                      skip = 2) %>% #skipping first two rows of .xls file
            column_to_rownames(var = "Sample_ID") %>% #assigning the sample names as the row names 
            mutate(month_collected = month(Date_Collected, 
                                           label = TRUE, 
                                           abbr = FALSE)) # creating streamlined column for graphing/analysis by date
head(coldata)

```

**Gene annotation data**

```{r}
# Read in gene annotations (GTF) file **However, this annotations file for the new genome doesn't really have useful information in it -- come back to this step once we create our manual annotation files**
gtf <- read.delim(
  "/work/clh162/OysterRNA24/hisat2_align/Cv_genome_RU_2025_shared/Cvi_RU25.gtf",
  comment.char = "#",
  header = FALSE,
  sep = "\t",
  quote = ""
)
# add in column names 
names(gtf) <- c(
  "seqname", "source", "feature", "start", "end",
  "score", "strand", "frame", "attribute"
)
```

#### Quick look at data

```{r}
# How many genes have zero counts across all samples?
sum(rowSums(counts_data) == 0) 
  # 1550 genes 

# How many genes have at least 10 counts total?
sum(rowSums(counts_data) >= 10) 
  # 25072 genes 
```

## DESeq2 Workflow

### 1. Initial data tidying

We are going to combine these two data sets off of the information that is shared between them: the **names** of our sequenced samples. Because of this, we need to make sure that the names of the samples are identical between these two data frames.

From looking at our two data frames uploaded in the last code block, the names of the sample in the `counts_data` have an additional suffix on the end of each. We should shorten these to match the names of the samples in the `coldata` frame.

```{r}
# rename the columns in the counts_data data frame to match sample names in the coldata data frame: getting rid of suffix starting with "_Gill" 
colnames(counts_data) <- sub("(24).*", "\\1", colnames(counts_data))
colnames(counts_data)

rownames(coldata) <- sub("(24).*", "\\1", rownames(coldata))
rownames(coldata)

# make sure all the row names in sample data and gene annotations matches all the column names in counts data (i.e. match on sample names)
all(rownames(coldata) %in% colnames(counts_data)) #TRUE
#all(rownames(rowranges) %in% rownames(counts_data)) # Run when annotations are added! 

# make sure these names are also in the same order 
all(rownames(coldata) == colnames(counts_data)) #FALSE
#all(rownames(rowranges) == rownames(counts_data)) # Run when annotations are added! 

# re-order 
counts_data[ ,rownames(coldata)] #this line works, so let's save it now
counts_data <- counts_data[ ,rownames(coldata)]
  # Alternative way to write to match on the order of the counts_data column sample names
  # coldata[colnames(counts_data), ] 
  # coldata <- coldata[colnames(counts_data), ]

# re-check that the names are in the same order
all(rownames(coldata) == colnames(counts_data)) #TRUE

```

Now that we have matching column/row names and they are in the right order, we can construct our SummarizedExperiment (se) and DESeqDataSet (dds) objects!

### 2. Create a se object

**SummarizedExperiment (se) object:** The idea of a se object is that it compiles all the gene expression data, gene features/annotations, and sample metadata into one cross-talking format. [The Bioconductor tutorial](https://carpentries-incubator.github.io/bioc-intro/60-next-steps.html#next-steps) talks about the value of creating a se object for inital omics data exploration. This tutorial summarizes how the se object is set up:

-   One (or more) assay(s) containing the quantitative omics data (expression data), stored as a matrix-like object. Features (genes, transcripts, proteins, …) are defined along the rows, and samples along the columns.
    -   The count object(s) will be saved in assays slot
-   A sample metadata slot containing sample information, stored as a data frame. Rows from this table represent samples (rows should be exact matches to the columns of the expression data which also hold the sample IDs).
    -   This sample information will be stored in `colData` slot (sample metadata)
-   A feature metadata slot containing information describing each gene, stored as a data frame. The rows of this data frame match exactly the rows of the expression data (i.e. the gene names).
    -   The rowranges object describing the genes will be stored in rowRanges slot (features metadata)

**Note:** We had to create our own annotation files using the `Diamond` program. Therefore, creating a summarized experiment object (se) was initially skipped, but returned to after annotation files were created.

```{r}
se <- SummarizedExperiment(
    assays = list(counts = as.matrix(counts_data)),
    colData = coldata,
    rowData = rowranges #make sure name matches actual annotation file 
)
head(se)
```

Now we can look at each part of our `se` object by calling the name of each section.

```{r}
head(assay(se))
head(rowData(se))
head(colData(se))
```

Let's save our `se` object for future use. We can do this using the `saveRDS()` function from baseR. If we want to load it in the future, we can use the `readRDS()` function also included in baseR. We can also save this object in `.csv` format if we want to send it to someone to use/open outside of R.

```{r}
saveRDS(se, file = "OysterRNA24_summarized_experiment.rds")
```

The same Bioconductor tutorial referenced above eventually creates a **DESeq2** data object from the **se** object, however, you can also create a **DESeq2** object directly from a gene count matrix.

### 3. Create a DESeq2 dataset object

Before creating our DESeq2 dataset object, let's make sure that the variables we want to run the analysis by are all factors + ordered in the way we want (i.e. the right reference level is selected). DESeq2 will not accept an **ordered** factor directly into its design formula, so we will check to make sure we only have regular factors.

```{r}
# Location 
coldata$Location <- factor(coldata$Location, levels = c("CMT","DAF"), ordered = FALSE) #reference level provided first: CMT
is.ordered(coldata$Location) 
  # FALSE
# checking explicitly to see what this column is reading as 
class(coldata$Location) 
  # factor 

# Month 
coldata$month_collected <- factor(coldata$month_collected, levels = c("June", "July", "August"), ordered = FALSE)
is.ordered(coldata$month_collected) 
  # FALSE
# checking explicitly to see what this column is reading as
class(coldata$month_collected) 
  # factor 
```

We will use the `DESeqDataSetFromMatrix()` command to create a dataset from our count data and sample data that is organized into the format required for the `DESeq()` command.

By comparing the dispersion estimates of different model designs our Data+ students settled on the following for their model design: design = \~ month \* location

This seemed to be because the interaction of month by site is different from each month to the next.

```{r}
dds <- DESeqDataSetFromMatrix(countData = counts_data,
                       colData = coldata,
                       design = ~ month_collected * Location) 
dds

# Save DESeq2 dataset object 
saveRDS(dds, file = "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/count_dds.rds")

# Reload DESeq2 dataset object 
dds <- readRDS("/work/clh162/OysterRNA24/DEG_analysis/DESeq2/count_dds.rds")
```

### 4. Perform Data quality assessment (i.e. quality control check)

[Source for this section:](https://github.com/hbctraining/DGE_workshop/blob/master/lessons/03_DGE_QC_analysis.md) Running a QC check is important to ensure that the results data is set up well for exploratory analysis. This step should be run either before or in parallel to conducting differential gene expression testing (running `DESeq()` on your **dds** object). Here, we will run it before! The two main tools used to do this are **visualization** and **sample clustering**.

The QC checks on the count data should include **sample-level** and **gene-level** steps to help us ensure that the samples/replicates look good.

### Sample-level QC analysis

First, we want to assess overall similarity between samples. This includes asking questions like:

-   Which samples are similar/different to each other?
-   Does this fit to the expectation from the experiment’s design?
-   What are the major sources of variation in the dataset?

To run sample-level QC, we will use **Principal Component Analysis (PCA)** and **hierarchical clustering** methods.

Our sample-level QC allows us to identify how well our replicates cluster together and observe whether our experimental condition represents the major source of variation in the data. Performing sample-level QC can also identify any sample outliers, which may need to be explored further to determine whether they need to be removed prior to DE analysis.

#### Transform data

It is common to use a **transformed format** of the count data for sample-level QC analysis and there are two main options [Source](https://nf-co.re/rnaseq/dev/docs/usage/differential_expression_analysis/de_rstudio): `variance-stabilised (vst)` or `regularised log-transformed (rlog)` counts. Because `vst` is the computationally faster of the two, it is usually preferred for use with larger datasets. Meanwhile, `rlog` is better able to withstand impacts of outliers.

Because we do not have many (any) outliers based on the super quick look we took at our results, let's use the `vst` option. Both of these transformation functions, `vst()` and `rlog()` have an argument blind, for whether the transformation should be blind to the sample information specified by the design formula. The default is `TRUE` and it should be used to compare samples in a manner wholly unbiased by the information about experimental groups (important to use this option for QC assessment).

```{r}
# transform dds into varianced-stabilized dataset (vsd)
vst <- vst(dds, blind = TRUE) # blind = TRUE is the default; leave as is! 

# Save transformed count data 
saveRDS(vst, file = "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/Data_QC_assessment/count_vst.rds")

# Load transformed count data 
vst_countdata <- readRDS("/work/clh162/OysterRNA24/DEG_analysis/DESeq2/Data_QC_assessment/count_vst.rds")
```

#### Visualization: PCA plot

The first figure that is helpful to make in QC analysis of the data is a principle component analysis (PCA) plot. `DESeq2` has a built in function to do this: `plotPCA()`, which will actually only accept a **transformed** dataset -- so it will throw an error if you try to pass it your `dds_results` object. In addition to the **transformed** dataset you need to provide a `ingroup` or the column in your sample data that you want to plot by.

```{r, error = TRUE}
# trying to run with dds_results object -- throws error. 
plotPCA(dds, intgroup = "Location")
```

The reason why we are creating a PCA plot is because we are trying to figure out how to account for the most variation in our count data. That way, we can know the best equation set-up for running our DESeq()

```{r}
# Reminder of the names of group/factors in our sample metadata for graphing  
colData(dds)
  # "Location"
  # "month_collected"

# using vst transformed data:
pca_vst_location <- plotPCA(vst, 
                        intgroup = "Location", 
                        ntop = 500) +
  labs(title = "PCA of variance-stabilised (vst) transformed counts by location")
pca_vst_location

pca_vst_month <- plotPCA(vst, 
                         intgroup = "month_collected", 
                         ntop = 500) +
  labs(title = "PCA of variance-stabilised (vst) transformed counts by month")
pca_vst_month

pca_vst_interaction <- plotPCA(vst, 
                               intgroup = c("Location", "month_collected"), 
                               ntop = 500) +
  labs(title = "PCA of variance-stabilised (vst) transformed counts by month*location")
pca_vst_interaction

```

To make that last PCA a little more clear, we can create our own annotated PCA plot:

```{r}
# Save data from interaction PCA plot
pca_vst_interaction <- plotPCA(vst, 
                               intgroup = c("Location", "month_collected"), 
                               ntop = 500,
                               returnData = TRUE)   # This argument tells R to return the PCA values 
# get % variance explained by PCA to include in PCA graph
percentVar <- attr(pca_vst_interaction, "percentVar")

# get a clean sample name to include in plot (biological replicate)
pca_vst_interaction$sample_id <- str_extract(pca_vst_interaction$name, "SJ\\d+")
pca_vst_interaction

# Create an annotated version of the interaction PCA
annotated_pca_plot <- ggplot(
  pca_vst_interaction,
  aes(x = PC1,
      y = PC2,
      # plot points so color is Location
      color = month_collected,
      # plot points so shape is month
      shape = Location)
) +
  geom_point(size = 3) +
  labs(title = "PCA of variance-stabilised transformed count data",
       subtitle = "Graph created for QC assessment pre-filtering of gene count data and running DESeq analysis",
       color = "Location",
       shape = "Month",
       x = paste0("PC1: ", round(percentVar[1] * 100), "% variance"),
       y = paste0("PC2: ", round(percentVar[2] * 100), "% variance")) +
  theme_bw()
annotated_pca_plot

    

# save the annotated plot 
ggsave("PCA_vst_annotated.png", plot = annotated_pca_plot, path = "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/Data_QC_assessment/")
```

#### Sample hierarchical clustering

The next thing we should do in our QC analysis is visualize the similarity between samples using hierarchical clustering. Hierarchical clustering involves grouping similar rows + columns of data to each other. So what we will do here is plot sample to sample distance.

The `DESeq2` package does not have its own function for hierarchical clustering but we can use the `pheatmap()` function from the `pheatmap` package (loaded above).

First, use the `dist()` function on the transformed count matrix to get sample-to-sample distances.

```{r}
# Pull the matrix of vst-transformed counts from the vst object
vst_mat <- assay(vst)  

### Compute pairwise correlation values
vst_cor <- cor(vst_mat)    ## cor() is a base R function

head(vst_cor)   ## check the output of cor(), make note of the rownames and colnames (should be the same)
```

Then visualize the sample-to-sample distances with a heatmap -- it'll give us an overview of similarities and dissimilarities between samples. You provide a hierarchical clustering to the heatmap function based on the sample distances, or else the heatmap function would calculate a clustering based on the distances between the rows/columns of the distance matrix.

```{r}
# Create a custom color palette for the heatmap -- using the RColorBrewer package (loaded above)
heat.colors <- colorRampPalette(rev(brewer.pal(9, "Greens")))(255)

# Create heatmap
hierarchical_clustering_plot <- pheatmap(vst_cor,
                                         color = heat.colors,
                                         fontsize_col = 8,
                                         fontsize_row = 8,
                                         main = "Heatmap of hierarchical clustering of variance-stabilised count data") 

# save the plot 
ggsave("hierarchical_clustering_vst_data.png", plot = hierarchical_clustering_plot, path = "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/Data_QC_assessment/")
```

### Gene-level QC analysis

#### Removing unexpresed genes

Before running differential expression analysis, it is good to omit genes that have little or no chance of being detected as differentially expressed. This will increase the power to detect differentially expressed genes.

The `DESeq2` program actually removes genes with `0` count across all samples, those with low mean normalized ocunts, and outliers automatically, but you definitely need to make sure to run this step if you are using a different differential expression program (for example: `EdgeR`) which doesn't automatically take care of this.

People will use different thresholds for this step, but let's remove all genes that lack counts greater than 10 in a minimum of 3 of our samples (this seems to be a very minimal threshold).

```{r}
nrow(dds) #28538 genes before filtering 

# Remove genes/rows that do not have > 10 counts in at least 3 (of the 18) samples   
Groupsize <- 3
keep <- rowSums(counts(dds) >= 10) >= Groupsize
head(keep) # Boolean of either T (the gene has more than 10 reads in at least 9/18 samples) or F (it doesn't)
dds_filtered <- dds[keep, ]
nrow(dds_filtered) 
  #21326 genes after filtering (got rid of ~7,000)

# save filtered dds object 
saveRDS(dds_filtered, file = "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/count_dds_filtered.rds")

# Reload DESeq2 dataset object 
# dds_filtered <- readRDS("/work/clh162/OysterRNA24/DEG_analysis/DESeq2/count_dds_filtered.rds")
```

### 5. Additional important pre-DESeq steps

#### Setting reference factor

We already did this above (before creating our DESeq object) but if you haven't already set your reference factor, make sure to do that before running `DESeq()`.

```{r}
dds_filtered$Location
dds_filtered$Location <- relevel(dds_filtered$Location, ref = "CMT")
# Another way to do this is with the factor() function - helpful if more than two levels for the variable you want to use as the factor  
#dds_filtered$Location <- factor(dds_filtered$Location, levels = c("CMT","DAF"))
```

#### Collapse technical replicates

If you have technical replicates (multiple samples collected from the same organism/timepoint), you need to fold them into one sample **before** running DESeq on the dds object. With this data, we don't have technical replicates, so we can move on to the next step.

**NOTE**: Be careful to not confuse technical replicates with biological replicates -- you should **NEVER** collapse biological replicates!

### 6. Run DESeq on dataset object

Now we are ready to run DESeq! Usually, people will save the results of DESeq back to the dds object name (to save space).

The `DESeq()` function is a wrapper function that simplifies the process of differential expression analysis by combining multiple steps into a single call [Source](https://nf-co.re/rnaseq/dev/docs/usage/differential_expression_analysis/de_rstudio). It makes the pipeline more user-friendly and ensures that all necessary pre-processing and statistical steps are executed in the correct order. As it runs, it will print statements about the steps as it is completing them. The key functions that DESeq2 calls include:

-   `estimateSizeFactors()`: to normalize the count data
    -   So NOTE that the normalization happens within running this one function! See this [source](https://github.com/hbctraining/DGE_workshop/blob/master/lessons/02_DGE_count_normalization.md) for more information about what is actually going on behind the scenes to calculate this normalization.
-   `estimateDispersion()`: to estimate the dispersion
-   `nbinomWaldTest()`: to perform differential expression test

```{r}
# Run DESeq

# on regular dds data object (28538 elements)
dds <- DESeq(dds)
# An experiment: running on trimmed data object (21326 elements)
dds_filtered <- DESeq(dds_filtered)

# for now subsequent work/analysis use the filtered dds set ! 
# save DESeqResults object of dds_filtered object 
saveRDS(dds_filtered, "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/DESeqResults_Objects/DESeqResults_dds_filtered.rds")

# load DESeqResults object of dds_filtered object 
dds_filtered <- readRDS("/work/clh162/OysterRNA24/DEG_analysis/DESeq2/DESeqResults_Objects/DESeqResults_dds_filtered.rds")
```

#### Model coefficient/comparison outputs

With our two reference factors being month (June, July, August) and Location (CMT, DAF) we have a resulting 6 coefficients in our output, which includes the intercept (or our two reference levels) and some of the interaction terms. However, the output will leave out a few comparisons, but we can manually call those to results if we need (see below) by listing **contrasts** for the `name =` function.

```{r}
# examine the available coefficients/comparisons from our DESeq model:
resultsNames(dds_filtered)
  # "Intercept" = The two reference levels: The average log expression for June at CMT 
  # "month_collected_July_vs_June" = Difference in average log expression between July and June at CMT 
  # "month_collected_August_vs_June" = Difference in average log expression between August and June at CMT
  # "Location_DAF_vs_CMT" = Difference in average log expression between DAF and CMT in June 
  # "month_collectedJuly.LocationDAF" = How much the July-vs-June change differs between DAF and CMT
  # "month_collectedAugust.LocationDAF" = How much the August-vs-June change differs between DAF and CMT

# How many samples fall into each category 
table(coldata$month_collected, coldata$Location)
```

#### Super quick look at general results

The `results()` function contains a number of arguments to customize the results table which is generated. You can read about these arguments by looking up `?results`. By saving the results of our DESeq run, we can see the reported results values for each of our genes.

We can use the `summary()` function to get a general idea of up and down regulated genes in our samples.

```{r}
# Save the general results of DESeq 
results <- results(dds_filtered)
# Order the results w/smallest padj values 
results <- results[order(results$padj), ]
head(results)
results
# default 0.1 padj-based results 
summary(results)
```

Looking at the summary of our DESeq results, we can see that the `results()` function uses a default padj value of `0.1`. We can change this by designating an alpha level in creating our results object: `alpha = #`.

**Q: How does changing the alpha level influence our DESeq2 objects / dataframes that result?**

If we call these two results objects, them in the `summary()` function, graph their MA plots, etc., they will **not** look different. The only time this difference will become apparent is when reading the adjusted p-values. This is because no genes (or in the graphs "points") are being removed from the results file, only the way they are being regarded as significant is changing.

Setting the `alpha =` to something other than the default value of `0.1` in the `results()` function changes how the independent filtering is performed: the value it is set to becomes the significance cutoff for the independent filtering, which changes the adjusted p-values themselves.

Whatever adjusted p-value cutoff (FDR) value you want to use, you should set `alpha` to that in your `results()` function. The `summary()` function will use the same value of `alpha` specified in the `results()` function and alternatively use the default of `alpha=0.1` if a `alpha` value is not specified in the `results()` function.

```{r}
# 0.05 padj-based results 
results0.05 <- results(dds_filtered, alpha = 0.05)
# Order the 0.05 padj-results w/smallest padj values 
results0.05 <- results0.05[order(results0.05$padj), ]
results0.05
summary(results0.05)
```

It may be interesting right off the bat to quantify the number of differentially expressed genes we have in our results dataframe based on the significance parameters we decide to use. Some standards that are commonly used include:

-   padj-value \< 0.05
-   logfoldchange \> 1 (or a Fold Change of 2)

**Quick sidebar** -- understanding fold change (FC) and log2foldchange (logFC)

log2(FC) = logFC FC = 2\^logFC

```{r}
library(data.table)
FC_vs_logFC_example <- data.table(
  Gene = c("Gene A", "Gene B", "Gene C"),
  Control_expression = c("100", "100", "200", "100"),
  Treatment_expression = c("200", "400", "100", "100"),
  Fold_change = c("2", "4", "0.5", "1"),
  log_Fold_change = c("1", "2", "-1", "0")
)
FC_vs_logFC_example

```

If there is exactly no difference in expression between groups, we would expect a FC of 1 or a logFC of 0.

We can hard set the significance cut off standards for reproducibility:

```{r}
### Set thresholds
padj.cutoff <- 0.05
logFC.cutoff <- 2 
```

The `logFC.cutoff` is set to 2; Remember, we are working with log2foldchanges so a logFC cut off of 2 translates to an actual fold change of 4 which seems more stringent.

We can see how these standards change the total \# of genes in a given results comparison that register as significantly differentially expressed.

```{r}
# all genes
nrow(as.data.frame(results0.05))
# only genes with padj <0.05
nrow(filter(as.data.frame(results0.05), padj<padj.cutoff))
# only genes with padj <0.05 and an absolute fold change > 4 (logFC > 2)
nrow(filter(as.data.frame(results0.05),padj<padj.cutoff,abs(log2FoldChange)>lfc.cutoff))
```

We can also plot dispersion estimates

```{r}
plotDispEsts(dds_filtered, 
             main = "Dispersion estimates for model design: ~ month * location") # `main = ` adding title 
```

#### Running interested pairwise-result comparisons

**Important NOTE:** the above created `results0.05` DESeqResults object is NOT answering the question of "how are all 21,326 genes in our dataset differentially expressed." It isn't doing this for two reasons:

1. We did not specify a `contrast = ` in creating the results object with the `results()` function. So it automatically (by default) used the last coefficient of the model the create that results object. 

2. Our model design (month * location) doesn't actually provide us a coefficient to ask that question! So our analysis should be comparing between the sites in a single month or between different months within a site (our model coefficients will give us this). 

To pull the results from a more specific comparison, we can use `name =` within the `results()` function with the name of the coefficient/comparison (see code chunk above). This information is stored as a `DESeqResults` object, which we can investigate with the `summary()` function so to get the actual data, we need to convert it to a dataframe. The pairwise-results that we are interested in (differences between sites at a given timepoint and differences over time at a single site) can be seen at the [Analysis section](#full-deg-analysis-for-oyster-rna-2024-data) below.

#### Checking our contrasts

If we had more than levels within our factor chosen to perform the DESeq by (here "Location"), we could also change which levels were being used for a direct comparison between these levels. An example is provided below.

```{r}
# resultsNames(dds_results_Location) #our design factor is Location
# example scenario where we have a third Location (PHI) that we want to compare to our reference level (CMT)
  # results(dds_results_Location, contrast = c("Location", "PHI", "CMT"))
```

### 7. Visualizing differential gene expression results

As a reminder, in order to test for differential expression, we operate on **raw counts** and use discrete distributions. So to plot DGE data, we should are using raw counts that are the output of our `DESeq()` result object.

#### MA Plot

A good plot to start with experimental analysis of DEG is MA Plots. MA Plots are a scatter plot of log fold change against the mean of normalized counts. It can help us visualize the genes that are being differentially expressed. The blue points are the genes that are significantly differentially expressed.

The command to do this from the `DESeq2` package is `plotMA()`.

The direction of the triangle shows the direction of the logfold change. Genes in the upper right and lower right corners would have high mean of normalized counts as well as extreme (either higher or lower) logfold changes and would be interesting genes to follow up on.

```{r}
# default 0.1 padj-based results
plotMA(results)
# 0.05 padj-based results 
plotMA(results0.05)
```

#### Volcano Plot

```{r}
# manually creating a volcano plot for default 0.1 padj-based results
results_df <- as.data.frame(results) %>% 
  rownames_to_column("gene_id") %>% 
  mutate(
    significance = ifelse(!is.na(padj) & padj < 0.05 & abs(log2FoldChange) >= 1,
                         "Significant", "Not significant")
  )

ggplot(results_df, aes(x = log2FoldChange, y = -log10(padj), color = significance)) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_vline(xintercept = c(-1, 1),
             linetype = "dashed") +
  scale_color_manual(values = c("red", "blue")) + 
  theme_bw() +
  labs(
    title = "DESeq2 results",
    x = "log2 fold change",
    y = "-log10 adjusted p-value",
    color = "Result"
  )
```

#### PCA Plot

Again, like we saw in the QC above, if you want to show your gene expression data in a PCA plot set up, it is best to do it with the transformed format of the data (in this case we have variance-stabilized transformed data).

**NOTE: That the PCA create here displays the same trends/variation distribution as the PCA run on the gene count data before filtering and DESeq analysis -- only need to run once!**

```{r}
# transform DESeq dds result object into varianced-stabilized dataset (vsd)
vst <- vst(dds_filtered, blind = TRUE) # blind = TRUE is the default; leave as is! 

pcaData <- plotPCA(vst, intgroup = c("Location","month_collected"), ntop = 500, returnData = TRUE)
# get % variance explained by PCA to include in PCA graph
percentVar <- attr(pcaData, "percentVar")

DESeq_pca <- ggplot(pcaData, 
       aes(x = PC1, y = PC2, color = month_collected, shape = Location, label = name)) + 
  geom_point(size = 3) + 
  #geom_text(vjust = -1, size = 2) +
  scale_color_manual(values = c(
    "June" = "lightblue", 
    "July" = "lightblue4",
    "August" = "blue4")) +
  labs(
    title = "PCA of RNA-seq Samples post DESeq",
    subtitle = "Based on top 500 most variable genes",
    x = paste0("PC1: ", round(percentVar[1] * 100), "% variance"),
    y = paste0("PC2: ", round(percentVar[2] * 100), "% variance"),
    color = "Month Collected",
    shape = "Location"
  ) + 
        theme_bw()

# save PCA plot of DESeq results 
ggsave(
    filename = "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/PCA_DESeqresults.png",
    plot = DESeq_pca, 
    width = 10, 
    height = 8, 
    dpi = 300)
```

#### Heatmap of Gene expression 

Still using our variance stabilized transformation DESeq results data. First, let's create a heatmap that displays our samples in the order of greatest similarly to each other (on the dendrogram that is created with the `Heatmap()` function).

```{r}
# Order DESeq results by smallest padj values 
results0.05 <- results0.05[order(results0.05$padj), ]
# get the top 50 genes based on adjusted p-value (50 lowest p-values = 50 most statistically significant genes)
top_genes <- rownames(results0.05)[1:500] 

mat <- assay(vst)[top_genes, ] # get expression values
mat <- t(scale(t(mat))) # scale each gene
heatmap_annotation <- HeatmapAnnotation(Location = colData(vst)$Location, 
                                        Month = colData(vst)$month_collected,
                                        # set the colors for both annotation groups 
                                        col = list(
                                          Location = c(
                                            "CMT" = "green3",
                                            "DAF" = "darkgreen"
                                            ),
                                          Month = c(
                                            "June" = "lightblue", 
                                            "July" = "lightblue4",
                                            "August" = "blue4"
                                            )
                                          )
                                        )

colData(vst)$month_collected <- factor(
    colData(vst)$month_collected,
    levels = c("June", "July", "August")
)


# create heatmap
heatmapDESeqresults <- Heatmap(mat,
        name = "Expression",
        # rect_gp = gpar(col = "white", lwd = 2), # if you want a border around each expression cell 
        top_annotation = heatmap_annotation,
        show_row_names = FALSE, # can change if want to see gene ids
        cluster_rows = TRUE,
        cluster_columns = TRUE)
heatmapDESeqresults
# need to figure out how to save this plot -- the following option doesn't work:
# outfile <- "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/heatmap_DESeqresults.pdf"
# 
# pdf(outfile,
#     width = 8,
#     height = 6)
# 
# draw(heatmapDESeqresults)
# 
# dev.off()
# 
# file.exists(outfile)
```

Let's create a second heatmap that plots our tissue samples in a more organized order (in sequential dates for each location)

```{r}
# create custom sample order 
sample_order <- order(colData(vst)$Location,
                      colData(vst)$month_collected)

# create heatmap annotation
heatmap_annotation_month <- HeatmapAnnotation(Location = colData(vst)$Location[sample_order],
                                              Month = colData(vst)$month_collected[sample_order],
                                        # set the colors for both annotation groups 
                                        col = list(
                                          Location = c(
                                            "CMT" = "green3",
                                            "DAF" = "darkgreen"
                                            ),
                                          Month = c(
                                            "June" = "lightblue", 
                                            "July" = "lightblue4",
                                            "August" = "blue4"
                                            )
                                          )
                                        )

colData(vst)$month_collected <- factor(
    colData(vst)$month_collected,
    levels = c("June", "July", "August")
)

heatmapDESeqresults <- Heatmap(mat[, sample_order],
        name = "Expression",
        # rect_gp = gpar(col = "white", lwd = 2), # if you want a border around each expression cell 
        top_annotation = heatmap_annotation_month,
        show_row_names = FALSE, # can change if want to see gene ids
        show_column_names = FALSE,
        cluster_rows = TRUE,
        cluster_columns = FALSE)
heatmapDESeqresults
```


# Full DEG Analysis for Oyster RNA 2024 data {#full-deg-analysis-for-oyster-rna-2024-data}

## Overall comparison of Sites

DAF vs. CMT

\^ with our current model design, there is NO interaction term that allows us to look at how gene expression varies between sites because the model design (design = \~ month_collected \* Location) assumes the location effect can be different in June, July, and August.

**SOMETHING THAT WOULD BE GOOD TO TEST** -- A test to determine which model should really be used in the design of the dds object. Common workflow for this:

1.  Fit the additive model:

\~ month_collected + Location

```{r}

```


2.  Fit the interaction model:

\~ month_collected \* Location

```{r}

```


3.  Compare them with an LRT: dds_lrt \<- DESeq( dds, test = "LRT", reduced = \~ month_collected + Location )

res_lrt \<- results(dds_lrt)

\^ Where above....

Full model: \~ month_collected \* Location 

Reduced model: \~ month_collected + Location

The LRT tests whether the interaction terms explain significantly more variation than the additive model.

4.  Count significant interaction genes in the dds_lrt result\

-   0–10 genes with significant interaction → there's little evidence that the location effect changes over months.

-   Hundreds or thousands of genes with significant interaction → many genes behave differently over time at the two locations, so keeping the interaction model is justified.

-   If the interaction is not significant, then the simpler additive model is usually preferred because it estimates one overall location effect.

-   If the interaction is significant, then you generally report location effects within each month, rather than averaging across months.

```{r}

```

## Pair-wise comparsion of Sites + Months

**Gene expression analyses between sites at each month**

### 1. DAF vs. CMT in June

**Extracting results data**

For these pair-wise comparison results, we can include the alpha argument and setting it to 0.05. The alpha value is the significance cutoff used for optimizing the independent filtering (by default it is set to 0.1). If the adjusted p-value cutoff (FDR) will be a value other than 0.1 (for our final list of significant genes), alpha should be set to that value.

```{r}
# load DESeq(dds_filtered) object -- if needed 
# dds_filtered <- readRDS("/work/clh162/OysterRNA24/DEG_analysis/DESeq2/DESeq(dds_filtered)object.rds")

# use the resultsNames() function to return the coefficients of the model to subset results 
# resultsNames(dds_filtered)

# create sub-results object 
res_DAF_vs_CMT_June <- results(
  dds_filtered,
  alpha = 0.05,
  name = "Location_DAF_vs_CMT"
)

# Save DESeqResults sub-object
saveRDS(res_DAF_vs_CMT_June, "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/DESeqResults_Objects/DESeqResults_0.05_june.rds")

# see summary of results 
summary(res_DAF_vs_CMT_June)

class(res_DAF_vs_CMT_June)

# convert results to data frame format to save them + add annotations 
res_0.05_june_df <- as.data.frame(res_DAF_vs_CMT_June) %>% 
  rownames_to_column("gene_id") %>% 
  arrange(padj)
head(res_0.05_june_df)

# save results from data frame format 
write_csv(res_0.05_june_df,
          file = "res_0.05_DAF_vs_CMT_June",
          path = "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/res_0.05_DAF_vs_CMT_June")

# Reloading data 
# res_0.05_DAF_vs_CMT_June <- read_csv("/work/clh162/OysterRNA24/DEG_analysis/DESeq2/res_0.05_DAF_vs_CMT_June")
# summary(res_0.05_DAF_vs_CMT_June)

```

#### MA Plot

```{r}
plotMA(res_DAF_vs_CMT_June, main = "DAF vs. CMT - June", ylim = c(-5, 5))
```


#### Volcano Plot 

**Create a function to graphing volcano plot of DEG** 

In this function the default values for assigning "significance" to DEGs are: 

* padj \< 0.05 
* log2FoldChange \> 1

For labels to be assigned to genes in the plot, they have to be even more differentially expressed, with default values of:

* padj \< 1e-7
* log2FoldChange > 2

```{r}
# volcano plot function
plot_volcano <- function(res, title, padj_cutoff = 0.05, lfc_cutoff = 1, 
                         label_padj = 1e-7, label_lfc = 2) { # requires six inputs: 
  # (1) res: a results object from a DESeq object and 
  # (2) a title for the volcano plot 
  # (3) cutoff of padj value for significance (default is 0.05)
  # (4) cutoff of log2FoldChange for significance (default is 1)
  # (5) cutoff of padj value needed to label differentially expressed gene by identity
  # (6) cutoff of log2FoldChange value ALSO needed to label differentially expressed gene by identity

  # define the data used by the function from the results object (need to transform it into a dataframe object)
  df <- as.data.frame(res) %>%
    tibble::rownames_to_column("gene_id")

  # define significant genes
  df <- df %>%
    mutate(
      Significant = !is.na(padj) &
        padj < padj_cutoff &
        abs(log2FoldChange) > lfc_cutoff
    )
  
  # of all the genes that meet the significance value, we will add a label of their names to ones that are "super significant"
  label_genes <- df %>%
    filter(
      !is.na(padj),
      padj < label_padj,
      abs(log2FoldChange) > label_lfc
    )
  
  # Volcano plot
  ggplot(df,
         aes(x = log2FoldChange, 
             y = -log10(padj), 
             color = Significant)
         ) +
    geom_point(alpha = 0.7) +
    geom_vline(xintercept = c(-lfc_cutoff, lfc_cutoff),
               linetype = "dashed") +
    geom_hline(yintercept = -log10(padj_cutoff),
               linetype = "dashed") +
    scale_color_manual(
      values = c(
      "FALSE" = "grey70", 
      "TRUE" = "red")
                       ) +
    theme_classic() +
    labs(title = title,
         x = "log2 fold change",
         y = "-log10 adjusted p-value",
         color = "Significant"
          ) #+
    # ggrepel::geom_text_repel(
    #   data = label_genes,
    #   aes(label = gene_id),
    #   size = 3,
    #   max.overlaps = Inf,
    #   show.legend = FALSE
    #   )
}
```


```{r}
# create volcano plot using volcano plot function 
p <- plot_volcano(res_DAF_vs_CMT_June, "DESeq2 results: DAF vs CMT in June")

# save volcano plot 
ggsave(
    filename = "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/Volcanoplots/sites_june.png",
    plot = p, 
    width = 10, 
    height = 8, 
    dpi = 300)
```

### 2. DAF vs. CMT in July

```{r}
# seeing results equivalent of Location_DAF_vs_CMT_July = Difference in average log expression between DAF and CMT in July
res_DAF_vs_CMT_July <- results(
    dds_filtered,
    alpha = 0.05,
    contrast = list(
        c("Location_DAF_vs_CMT",
          "month_collectedJuly.LocationDAF")
    )
)

saveRDS(res_DAF_vs_CMT_July, "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/DESeqResults_Objects/DESeqResults_0.05_july.rds")

summary(res_DAF_vs_CMT_July)

res_july_df <- as.data.frame(res_DAF_vs_CMT_July) %>% 
  rownames_to_column("gene_id") %>% 
  arrange(padj)

write_csv(res_july_df,
          file = "res_DAF_vs_CMT_July",
          path = "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/res_0.05_DAF_vs_CMT_July")
```

#### MA Plot 

```{r}
plotMA(res_DAF_vs_CMT_July, main = "DAF vs. CMT - July", ylim = c(-5, 5))
```

#### Volcano Plot 

```{r}
p <- plot_volcano(res_DAF_vs_CMT_July, "DESeq2 results: DAF vs CMT in July")
ggsave(
    filename = "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/Volcanoplots/sites_july.png",
    plot = p, 
    width = 10, 
    height = 8, 
    dpi = 300)
```


### 3. DAF vs. CMT in August

```{r}
# seeing results equivalent of Location_DAF_vs_CMT_August = Difference in average log expression between DAF and CMT in August 
res_DAF_vs_CMT_August <- results(
    dds_filtered,
    alpha = 0.05,
    contrast = list(
        c("Location_DAF_vs_CMT",
          "month_collectedAugust.LocationDAF")
    )
)

saveRDS(res_DAF_vs_CMT_August, "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/DESeqResults_Objects/DESeqResults_0.05_august.rds")

summary(res_DAF_vs_CMT_August)

res_aug_df <- as.data.frame(res_DAF_vs_CMT_August) %>% 
  rownames_to_column("gene_id") %>% 
  arrange(padj)

write_csv(res_aug_df,
          file = "res_DAF_vs_CMT_August",
          path = "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/res_0.05_DAF_vs_CMT_August")
```

#### MA Plot 

```{r}
plotMA(res_DAF_vs_CMT_August, main = "DAF vs. CMT - August", ylim = c(-5, 5))
```

#### Volcano Plot 

```{r}
p <- plot_volcano(res_DAF_vs_CMT_August, "DESeq2 results: DAF vs CMT in August")
ggsave(
    filename = "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/Volcanoplots/sites_august.png",
    plot = p, 
    width = 10, 
    height = 8, 
    dpi = 300)
```


**Gene expression analyses over time (between months) at each site**

### 4. DAF July to June

```{r}
res_DAF_July_vs_June <- results(
    dds_filtered,
    alpha = 0.05,
    contrast = list(
        c("month_collected_July_vs_June",
          "month_collectedJuly.LocationDAF")
    )
)

summary(res_DAF_July_vs_June)

res_DAF_July_vs_June_df <- as.data.frame(res_DAF_July_vs_June) %>% 
  rownames_to_column("gene_id") %>% 
  arrange(padj)
res_DAF_July_vs_June_df

write_csv(res_DAF_July_vs_June_df,
          file = "res_DAF_July_vs_June",
          path = "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/res_0.05_July_vs_June_DAF")
```

#### MA Plot 

```{r}
plotMA(res_DAF_July_vs_June, main = "DAF: July vs. June", ylim = c(-5, 5))
```

#### Volcano Plot 

```{r}
p <- plot_volcano(res_DAF_July_vs_June, "DESeq2 results: July vs. June @ DAF")
ggsave(
    filename = "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/Volcanoplots/daf_july_vs_june.png",
    plot = p, 
    width = 10, 
    height = 8, 
    dpi = 300)
```

### 5. DAF August to June

```{r}
res_DAF_August_vs_June <- results(
    dds_filtered,
    alpha = 0.05,
    contrast = list(
        c("month_collected_August_vs_June",
          "month_collectedAugust.LocationDAF")
    )
)

summary(res_DAF_August_vs_June)

res_DAF_August_vs_June_df <- as.data.frame(res_DAF_August_vs_June) %>% 
  rownames_to_column("gene_id") %>% 
  arrange(padj)

write_csv(res_DAF_August_vs_June_df,
          file = "res_DAF_August_vs_June",
          path = "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/res_0.05_August_vs_June_DAF")
```

#### MA Plot 

```{r}
plotMA(res_DAF_August_vs_June, main = "DAF: August vs. June", ylim = c(-5, 5))
```

#### Volcano Plot 

```{r}
p <- plot_volcano(res_DAF_August_vs_June, "DESeq2 results: August vs. June @ DAF")
ggsave(
    filename = "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/Volcanoplots/daf_august_vs_june.png",
    plot = p, 
    width = 10, 
    height = 8, 
    dpi = 300)
```


### 6. DAF August to July

\^ This one isn't explicity outlined in coefficients from the model. What we need to get to it is....

(August vs June at DAF) minus (July vs June at DAF)

which expands to:

month_collected_August_vs_June + month_collectedAugust.LocationDAF - month_collected_July_vs_June - month_collectedJuly.LocationDAF

```{r}
res_DAF_August_vs_July <- results(
    dds_filtered,
    alpha = 0.05,
    contrast = list(
        c("month_collected_August_vs_June", #Aug vs. June @ CMT
          "month_collectedAugust.LocationDAF"), #August-vs-June change differs between DAF and CMT
        c("month_collected_July_vs_June", #July vs. June @ CMT 
          "month_collectedJuly.LocationDAF") #July-vs-June change differs between DAF and CMT
    )
)

summary(res_DAF_August_vs_July)

res_DAF_August_vs_July_df <- as.data.frame(res_DAF_August_vs_July) %>% 
  rownames_to_column("gene_id") %>% 
  arrange(padj)

write_csv(res_DAF_August_vs_July_df,
          file = "res_DAF_August_vs_July",
          path = "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/res_0.05_August_vs_July_DAF")
```

#### MA Plot 

```{r}
plotMA(res_DAF_August_vs_July, main = "DAF: August vs. July", ylim = c(-5, 5))
```

#### Volcano Plot 

```{r}
p <- plot_volcano(res_DAF_August_vs_July, "DESeq2 results: August vs. July @ DAF")
ggsave(
    filename = "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/Volcanoplots/daf_august_vs_july.png",
    plot = p, 
    width = 10, 
    height = 8, 
    dpi = 300)
```

### 7. CMT July to June

```{r}
res_CMT_July_vs_June <- results(
    dds_filtered,
    alpha = 0.05,
    name = "month_collected_July_vs_June"
    )

summary(res_CMT_July_vs_June)

res_CMT_July_vs_June_df <- as.data.frame(res_CMT_July_vs_June) %>% 
  rownames_to_column("gene_id") %>% 
  arrange(padj)
res_CMT_July_vs_June_df

write_csv(res_CMT_July_vs_June_df,
          file = "res_CMT_July_vs_June",
          path = "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/res_0.05_July_vs_June_CMT")
```

#### MA Plot 

```{r}
plotMA(res_CMT_July_vs_June, main = "CMT: July vs. June", ylim = c(-5, 5))
```

#### Volcano Plot 

```{r}
p <- plot_volcano(res_CMT_July_vs_June, "DESeq2 results: July vs. June @ CMT")
ggsave(
    filename = "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/Volcanoplots/cmt_july_vs_june.png",
    plot = p, 
    width = 10, 
    height = 8, 
    dpi = 300)
```

### 8. CMT August to June

```{r}
res_CMT_August_vs_June <- results(
    dds_filtered,
    alpha = 0.05,
    name = "month_collected_August_vs_June"
    )

summary(res_CMT_August_vs_June)

res_CMT_August_vs_June_df <- as.data.frame(res_CMT_August_vs_June) %>% 
  rownames_to_column("gene_id") %>% 
  arrange(padj)

write_csv(res_CMT_August_vs_June_df,
          file = "res_CMT_August_vs_June",
          path = "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/res_0.05_August_vs_June_CMT")
```

#### MA Plot 

```{r}
plotMA(res_CMT_August_vs_June, main = "CMT: August vs. June", ylim = c(-5, 5))
```

#### Volcano Plot 

```{r}
p <- plot_volcano(res_CMT_August_vs_June, "DESeq2 results: August vs. June @ CMT")
ggsave(
    filename = "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/Volcanoplots/cmt_august_vs_june.png",
    plot = p, 
    width = 10, 
    height = 8, 
    dpi = 300)
```

### 9. CMT August to July

```{r}
res_CMT_August_vs_July <- results(
    dds_filtered,
    alpha = 0.05,
    contrast = list(
        "month_collected_August_vs_June",
        "month_collected_July_vs_June"
    )
)

summary(res_CMT_August_vs_July)

res_CMT_August_vs_July_df <- as.data.frame(res_CMT_August_vs_July) %>% 
  rownames_to_column("gene_id") %>% 
  arrange(padj)

write_csv(res_CMT_August_vs_July_df,
          file = "res_CMT_August_vs_July",
          path = "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/res_0.05_August_vs_July_CMT")
```

#### MA Plot 

```{r}
plotMA(res_CMT_August_vs_July, main = "CMT: August vs. July", ylim = c(-5, 5))
```

#### Volcano Plot 

```{r}
p <- plot_volcano(res_CMT_August_vs_July, "DESeq2 results: August vs. July @ CMT")
ggsave(
    filename = "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/Volcanoplots/cmt_august_vs_july.png",
    plot = p, 
    width = 10, 
    height = 8, 
    dpi = 300)
```

## Summary analyses of differentially expressed genes between the sites across these timepoints

Using guidelines of **pvalues < 0.05** and **LFC > 1** as our cutoff for significance. The adjusted p-value limits the expected false discovery rate, while the fold-change threshold excludes effects below the chosen reporting minimum (in this case, less than 2-fold change). Based on these cutoff values, how many of our DEGs (n = 21,326) are overall significantly differentially expressed?

Again, in the context of this workflow, our model design (`~ month * location`) doesn't actually provide us a coefficient to ask that question! So, our analysis should be comparing between the sites in a single month or between different months within a site (our model coefficients will give us this). 

```{r}
# reload DESeqResults object for filtered count data if not already loaded 
# results0.05 <- readRDS("/work/clh162/OysterRNA24/DEG_analysis/DESeq2/DESeqResults_Objects")

# count significant DEGs 
sum(results0.05$padj < 0.05 & 
      abs(results0.05$log2FoldChange) > 1, na.rm = TRUE) # 172 DEGs

# Generate lists of total significant upregulated and downregulated gene ids
# upregulated DEGs
upregulated_DEG <- subset(results0.05, padj < 0.05 & log2FoldChange > 1)
upregulated_DEG
upregulated_DEG_ids <- rownames(upregulated_DEG)
upregulated_DEG_ids
# downregulated DEGs
downregulated_DEG <- subset(results0.05, padj < 0.05 & log2FoldChange < 1)
downregulated_DEG_ids <- rownames(downregulated_DEG)
downregulated_DEG_ids
```

1. How many are significantly differentially expressed each month? Are the same significant genes significant across the months?  

```{r}
# reload DESeqResults objects for month-based analysis if not already loaded
res_DAF_vs_CMT_June <- readRDS("/work/clh162/OysterRNA24/DEG_analysis/DESeq2/DESeqResults_Objects/DESeqResults_0.05_june.rds")
res_DAF_vs_CMT_July <- readRDS("/work/clh162/OysterRNA24/DEG_analysis/DESeq2/DESeqResults_Objects/DESeqResults_0.05_july.rds")
res_DAF_vs_CMT_August <- readRDS("/work/clh162/OysterRNA24/DEG_analysis/DESeq2/DESeqResults_Objects/DESeqResults_0.05_august.rds")

# count significant DEGs per month (shows how much divergence exists between locations each month)
sum(res_DAF_vs_CMT_June$padj < 0.05 & 
      abs(res_DAF_vs_CMT_June$log2FoldChange) > 1, na.rm = TRUE) # 141 DEG in June  
sum(res_DAF_vs_CMT_July$padj < 0.05 & 
      abs(res_DAF_vs_CMT_July$log2FoldChange) > 1, na.rm = TRUE) # 109 DEG in July
sum(res_DAF_vs_CMT_August$padj < 0.05 & 
      abs(res_DAF_vs_CMT_August$log2FoldChange) > 1, na.rm = TRUE) # 253 DEG in August  

# determine if the different significant genes are consistent each month
# get list of significant DEGs ids each month 
sig_june <- rownames(subset(
  res_DAF_vs_CMT_June,
  padj < 0.05 & 
    abs(log2FoldChange) > 1
    )
  ) # 141

sig_july <- rownames(subset(
  res_DAF_vs_CMT_July,
  padj < 0.05 & 
    abs(log2FoldChange) > 1
    )
  ) # 109

sig_august <- rownames(subset(
  res_DAF_vs_CMT_August,
  padj < 0.05 & 
    abs(log2FoldChange) > 1
    ) 
  ) # 253
  
# count genes that are consistently significant across all three months
length(Reduce(intersect,
              list(sig_june, sig_july, sig_august))) # 9 genes 
# identity of genes that are consistently significant across all three months 
Reduce(intersect,
              list(sig_june, sig_july, sig_august))
# count genes that are consistently significant across pairwise comparison of months 
length(intersect(sig_june, sig_july)) # 16 genes 
length(intersect(sig_june, sig_august)) # 36 genes 
length(intersect(sig_july, sig_august)) # 33 genes 
```

There were 141, 109, and 253 significantly differentially expressed genes (DEGs) between the two sites in June, July, and August respectively. Only nine of these genes are consistently significantly differentially expressed across all three months ("evm.TU.Chr05.304"  "evm.TU.Chr05.1151" "evm.TU.Chr02.671"  "evm.TU.Chr02.1009" "evm.TU.Chr02.1303" "evm.TU.Chr02.1304" "evm.TU.Chr02.1319" "evm.TU.Chr02.1320" "evm.TU.Chr02.1794").

2. What about the differentiation of significantly upregulated and downregulated genes between the months. Again, is the identity of these genes similar between months at these timepoints? 

```{r}
# count significant DEGs 

# Generate (1) result objects, (2) gene ID lists, and (3) counts of total significant upregulated and downregulated DEGs for each month 
# June
sig_up_june <- subset(res_DAF_vs_CMT_June, padj < 0.05 & log2FoldChange > 1) # 67 of 141
sig_up_june_ids <- rownames(sig_up_june)
sig_up_june_count <- sum(
  res_DAF_vs_CMT_June$padj < 0.05 &
    res_DAF_vs_CMT_June$log2FoldChange > 1,
  na.rm = TRUE
)

sig_down_june <- subset(res_DAF_vs_CMT_June, padj < 0.05 & log2FoldChange < -1) # 74 of 141
sig_down_june_ids <- rownames(sig_down_june)
sig_down_june_count <- sum(
  res_DAF_vs_CMT_June$padj < 0.05 &
    res_DAF_vs_CMT_June$log2FoldChange < -1,
  na.rm = TRUE
)

# July 
sig_up_july <- subset(res_DAF_vs_CMT_July, padj < 0.05 & log2FoldChange > 1) # 50 of 109
sig_up_july_ids <- rownames(sig_up_july)
sig_up_july_count <- sum(
  res_DAF_vs_CMT_July$padj < 0.05 &
    res_DAF_vs_CMT_July$log2FoldChange > 1,
  na.rm = TRUE
)

sig_down_july <- subset(res_DAF_vs_CMT_July, padj < 0.05 & log2FoldChange < -1) # 59 of 109 
sig_down_july_ids <- rownames(sig_down_july)
sig_down_july_count <- sum(
  res_DAF_vs_CMT_July$padj < 0.05 &
    res_DAF_vs_CMT_July$log2FoldChange < -1,
  na.rm = TRUE
)

# August 
sig_up_august <- subset(res_DAF_vs_CMT_August, padj < 0.05 & log2FoldChange > 1) # 111 of 253
sig_up_august_ids <- rownames(sig_up_august)
sig_up_august_count <- sum(
  res_DAF_vs_CMT_August$padj < 0.05 &
    res_DAF_vs_CMT_August$log2FoldChange > 1,
  na.rm = TRUE
)

sig_down_august <- subset(res_DAF_vs_CMT_August, padj < 0.05 & log2FoldChange < -1) # 142 of 253 
sig_down_august_ids <- rownames(sig_down_august)
sig_down_august_count <- sum(
  res_DAF_vs_CMT_August$padj < 0.05 &
    res_DAF_vs_CMT_August$log2FoldChange < -1,
  na.rm = TRUE
)
```

Let's save the totals of these significantly up and downregulated DEGs between months into a table format:

```{r}
# put counts into table
sig_DEGs_count_table <- data.frame(
  Month = c("June", "July", "August"),
  Up_in_CMT = c(sig_up_june_count, sig_up_july_count, sig_up_august_count),
  Up_in_DAF = c(sig_down_june_count, sig_down_july_count, sig_down_august_count)
)
sig_DEGs_count_table

# save table 
write.csv(
  sig_DEGs_count_table,
  file = "/work/clh162/OysterRNA24/DEG_analysis/DESeq2/sig_DEGs_count_table.csv",
  row.names = FALSE
)
```


# Troubleshooting code

```{r}
# TEST 7/1 -- using result() set with alpha = 0.1 and DAF as reference site through goseq() to compare with Sophie

# creating new dds object to relevel the reference site to be DAF 
dds_filtered_DAF <- dds_filtered
dds_filtered_DAF$Location <- relevel(dds_filtered$Location, ref = "DAF")
dds_filtered_DAF$Location

# Running DESeq()
dds_filtered_DAF <- DESeq(dds_filtered_DAF)

# finding names of coefficients from DESeq model 
resultsNames(dds_filtered_DAF)

# create sub-results object 
dds_filtered_DAF <- results(
  dds_filtered_DAF,
  alpha = 0.1,
  name = "Location_CMT_vs_DAF"
)

# see summary of results 
summary(dds_filtered_DAF)

# convert results to data frame format to save them + add annotations 
res_0.1_CMT_vs_DAF_June <- as.data.frame(dds_filtered_DAF) %>% 
  rownames_to_column("gene_id") %>% 
  arrange(padj)

# load old to new gene ID file for ENSEMBL ID communication
RU17_RU25_gene_IDs <- read_csv("/work/clh162/OysterRNA24/DEG_analysis/GO/RU17_RU25_gene_IDs")

# run a left_join() to get all the ENSEMBL gene names added to all the genes in this results set 
res_0.1_CMT_vs_DAF_June <- left_join(x = res_0.1_CMT_vs_DAF_June,
          y = RU17_RU25_gene_IDs,
          by = join_by(gene_id == RU25_gene)) # column names for RU25 genes are different between the two data sets 
res_0.1_CMT_vs_DAF_June

# Create DE gene vector where 1 = significant and 0 = not
DE_genes_vector <- as.integer(
  res_0.1_CMT_vs_DAF_June$padj < 0.05 &
  !is.na(res_0.1_CMT_vs_DAF_June$padj) # exclude DE genes that do not have a padj value 
)

# add ensembl_gene_id names to the lengths data 
names(DE_genes_vector) <- res_0.1_CMT_vs_DAF_June$gene

# Rest of goseq() pipeline was carried out using already written code in 06_GO_analysis.qmd
```

