---
title: "OysterRNA24 DESeq2 Analysis"
author: "Callie Hundley"
format: 
  html:
    df-print: paged 
    code-fold: true 
    code-tools: true
editor: visual
---

# Project Overview 


## Loading packages 

```{r}
library(tidyverse)
library(lubridate)
library(readxl)
# BiocManager::install("DESeq2") # <- install only once!
library(DESeq2)
# install.packages("pheatmap")
library(pheatmap)
library(RColorBrewer)
library(ggplot2)
```


## Loading data 

```{r}
# Read in the gene count data 
counts_data <- read_tsv("/work/clh162/OysterRNA24/hisat2_align/featureCounts_counts/gene_counts_matrix.txt") %>% 
  column_to_rownames(var = "Geneid") #assigning the gene names as the row names 
head(counts_data)


# Read in the sample data (defining sequenced samples)
coldata <- read_excel("/work/clh162/OysterRNA24/xRNA for BC Summer 2024.xlsx", 
    skip = 2) %>% #skipping first two rows of .xls file
  column_to_rownames(var = "Sample_ID") %>% #assigning the sample names as the row names 
  mutate(Month_collected = month(Date_Collected, label = TRUE, abbr = FALSE)) # creating streamlined column for graphing/analysis by date
head(coldata)
```

## DESeq2 Workflow 

### 1. Initial data tidying 

We are going to combine these two data sets off of the information that is shared  between them: the **names** of our sequenced samples. Because of this, we need to make sure that the names of the samples are identical between these two data frames.

From looking at our two data frames uploaded in the last code block, the names of the sample in the `counts_data` have an additional suffix on the end of each. We should shorten these to match the names of the samples in the `coldata` frame. 


```{r}
# rename the columns in the counts_data data frame to match sample names in the coldata data frame 
colnames(counts_data) <- sub("(_Gill).*", "\\1", colnames(counts_data))

colnames(counts_data)
rownames(coldata)

# make sure all the row names in coldata matches all the column names in counts_data
all(rownames(coldata) %in% colnames(counts_data)) #TRUE

# make sure these names are also in the same order 
all(rownames(coldata) == colnames(counts_data)) #FALSE

# re-order 
counts_data[ ,rownames(coldata)]#this line works, so let's save it now
counts_data <- counts_data[ ,rownames(coldata)]
# Alternative way to write to match on the order of the counts_data column sample names
# coldata[colnames(counts_data), ] 
# coldata <- coldata[colnames(counts_data), ]

# re-check that the names are in the same order
all(rownames(coldata) == colnames(counts_data)) #TRUE

```

Now that we have matching column/row names and they are in the right order, we can construct our SummarizedExperiment (se) and DESeqDataSet (dds) objects! 

### 2. Create a se object 

**SummarizedExperiment (se) object:**
The idea of a se object is that it compiles all the gene expression data, gene features/annotations, and sample metadata into one cross-talking format. [The Bioconductor tutorial](https://carpentries-incubator.github.io/bioc-intro/60-next-steps.html#next-steps) talks about the value of creating a se object for inital omics data exploration. This tutorial summarizes how the se object is set up: 

* One (or more) assay(s) containing the quantitative omics data (expression data), stored as a matrix-like object. Features (genes, transcripts, proteins, …) are defined along the rows, and samples along the columns.
  * The count object(s) will be saved in assays slot

* A sample metadata slot containing sample information, stored as a data frame. Rows from this table represent samples (rows should be exact matches to the columns of the expression data which also hold the sample IDs). 
  * This sample information will be stored in `colData` slot (sample metadata)

* A feature metadata slot containing information describing each gene, stored as a data frame. The rows of this data frame match exactly the rows of the expression data (i.e. the gene names).
  * The rowranges object describing the genes will be stored in rowRanges slot (features metadata)


```{r}
se <- SummarizedExperiment(
    assays = list(counts = as.matrix(counts_data)),
    rowRanges = as(rowranges, "GRanges"),
    colData = coldata
)
```


The same Bioconductor tutorial referenced above eventually creates a **DESeq2** data object from the **se** object, however, you can also create a **DESeq2** object directly from a gene count matrix. 

### 2. Create a DESeq2 dataset object 

We will use the `DESeqDataSetFromMatrix()` command to create a dataset from our count data and sample data that is organized into the format required for the `DESeq()` command.  

```{r}
dds <- DESeqDataSetFromMatrix(countData = counts_data,
                       colData = coldata,
                       design = ~ Location) 
dds
```


### 3. Important pre-DESeq steps 

#### Removing unexpresed genes 

People will use different thresholds for this step, but let's remove genes that have fewer than 10 reads total (between all samples) (this seems to be a very minimal threshold). 

```{r}
nrow(dds) #28538 genes before filtering 

# Remove genes/rows that do not have > 10 counts in at least 9 (of the 18) samples   
Groupsize <- 3
keep <- rowSums(counts(dds) >= 10) >= Groupsize
head(keep) # Boolean of either T (the gene has more than 10 reads in at least 9/18 samples) or F (it doesn't)
dds_filtered <- dds[keep, ]
nrow(dds_filtered) #21326 genes after filtering (got rid of ~7,000)
```

#### Setting reference factor 

```{r}
dds_filtered$Location
dds_filtered$Location <- relevel(dds_filtered$Location, ref = "CMT")
# Another way to do this is with the factor() function - helpful if more than two levels for the variable you want to use as the factor  
#dds_filtered$Location <- factor(dds_filtered$Location, levels = c("CMT","DAF"))
```

#### Collapse technical replicates

If you have technical replicates (multiple samples collected from the same organism/timepoint), you need to fold them into one sample **before** running DESeq on the dds object. With this data, we don't have technical replicates, so we can move on to the next step. 

Be careful to not confuse technical replicates with biological replicates -- you should **NEVER** collapse biological replicates! 


### 4. Run DESeq on dataset object 

Now we are ready to run DESeq! Usually, people will save the results of DESeq back to the dds object name (to save space). Here, we aren't doing that because we have more space on the DCC and it is nice for reproducibility to practice these naming schemes. 

The `DESeq()` function is a wrapper function that simplifies the process of differential expression analysis by combining multiple steps into a single call [Source](https://nf-co.re/rnaseq/dev/docs/usage/differential_expression_analysis/de_rstudio). It makes the pipeline more user-friendly and ensures that all necessary pre-processing and statistical steps are executed in the correct order. The key functions that DESeq2 calls include:

* `estimateSizeFactors()`: to normalise the count data;
* `estimateDispersion()`: to estimate the dispersion;
* `nbinomWaldTest()`: to perform differential expression test.

```{r}
# Run DESeq
dds_results <- DESeq(dds_filtered)
```

#### Super quick look at results 

The `results()` function contains a number of arguments to customize the results table which is generated. You can read about these arguments by looking up `?results`. By saving the results of our DESeq run, we can see the reported results values for each of our genes. 

We can use the `summary()` function to get a general idea of up and down regulated genes in our samples.

```{r}
# Save the results of DESeq 
results <- results(dds_results)
results

# default 0.1 padj-based results 
summary(results)
```

Looking at the summary of our DESeq results, we can see that it uses a default padj value of `0.1`. We can change this by designating an alpha level in creating our results object: `alpha = #`.

```{r}
# 0.05 padj-based results 
results0.05 <- results(dds_results, alpha = 0.05)
summary(results0.05)
```


### 5. Perform Data quality assessment (i.e. quality control check)

Running a QC check is important to ensure that the results data is set up well for exploratory analysis. This step should be run either before or in parallel to conducting differential gene expression testing (see below). The two main tools used to do this are **visualization** and **sample clustering**. 

#### Transform data

It is common to use a transformed format of the count data for QC analysis and there are two main options [Source](https://nf-co.re/rnaseq/dev/docs/usage/differential_expression_analysis/de_rstudio): `variance-stabilised (vst)` or `regularised log-transformed (rlog)` counts. Because `vst` is the computationally faster of the two, it is usually  preferred for use with larger datasets. Meanwhile, `rlog` is better able to withstand impacts of outliers. 

Because we do not have many (any) outliers based on the super quick look we took at our results, let's use the `vst` option. Both of these transformation functions, `vst()` and `rlog()` have an argument blind, for whether the transformation should be blind to the sample information specified by the design formula. The default is `TRUE` and it should be used to compare samples in a manner wholly unbiased by the information about experimental groups.

```{r}
vst_results <- vst(dds_results, blind = TRUE)
```

#### Visualization: PCA plot 

The first figure that is helpful to make in QC analysis of the data is a principle component analysis (PCA) plot. The function to do this `plotPCA()` will actually only accept a **transformed** dataset -- so it will throw an error if you try to pass it your `dds_results` object.  

```{r, error = TRUE}
# trying to run with dds_results obejct -- throws error. 
plotPCA(dds_results, intgroup = "Location")
```

```{r}
# Reminder of the names of group/factors in our sample metadata for graphing  
colData(dds_results)

# using vst transformed data:
pca_plot_vst <- plotPCA(vst_results, intgroup = c("Location", "Month_collected"), ntop = 500) +
  labs(title = "PCA of variance-stabilised (vst) transformed DESeq count data")
pca_plot_vst

# save the plot 
ggsave("PCA_vst_data.png", plot = pca_plot_vst, path = "/work/clh162/OysterRNA24/DESeq2/Data_QC_assessment")
```

#### Sample clustering 

The next thing we should do in our QC analysis is visualize the similarity between samples using hierarchical clustering. Hierarchical clustering involves grouping similar rows + columns of data to each other. So what we will do here is plot sample to sample distance.

The `DESeq2` package does not have its own function for hierarchical clustering but we can use the `pheatmap()` function from the `pheatmap` package (loaded above). 

First, use the `dist()` function on the transformed count matrix to get sample-to-sample distances.

```{r}
# Pull the matrix of vst-transformed counts from the vst_results object
sampleDists <- dist(t(assay(vst_results)))  # Calculate pairwise distances between samples using the dist() function with Euclidean distance as the default method. By transposing the matrix with t(), we ensure that samples become rows and genes become columns, so that the dist function computes pairwise distances between samples.

# Convert distances to a matrix
sampleDistMatrix <- as.matrix(sampleDists)
```

Then visualize the sample-to-sample distances with a heatmap -- it'll give us an overview of similarities and dissimilarities between samples. You provide a hierarchical clustering to the heatmap function based on the sample distances, or else the heatmap function would calculate a clustering based on the distances between the rows/columns of the distance matrix.

```{r}
# Create a custom color palette for the heatmap -- using the RColorBrewer package (loaded above)
colors <- colorRampPalette(rev(brewer.pal(9, "Greens")))(255)

# Create heatmap
hierarchical_clustering_plot <- pheatmap(sampleDistMatrix,
                                         clustering_distance_rows = sampleDists,
                                         clustering_distance_cols = sampleDists,
                                         col = colors,
                                         fontsize_col = 8,
                                         fontsize_row = 8,
                                         main = "Heatmap of hierarchical clustering of variance-stabilised (vst) transformed DESeq count data") 

# save the plot 
ggsave("hierarchical_clustering_vst_data.png", plot = hierarchical_clustering_plot, path = "/work/clh162/OysterRNA24/DESeq2/Data_QC_assessment")
```


### 6. Further exploration of DESeq results (back to count data)





```{r}

```

#### Checking our contrasts

If we had more than levels within our factor chosen to perform the DESeq by (here "Location"), we could also change which levels were being used for a direct comparison between these levels. An example is provided below.

```{r}
resultsNames(dds_results) #our design factor is Location
# example scenario where we have a third Location (PHI) that we want to compare to our reference level (CMT)
  # results(dds_results, contrast = c("Location", "PHI", "CMT"))
```

### 7. Visualizing differential gene expression results 

As a reminder, in order to test for differential expression, we operate on **raw counts** and use discrete distributions. So to plot DGE data, we should are using raw counts that are the output of our `DESeq()` result object. 

#### MA Plot 

MA Plots are a scatterplot of log fold change against the mean of normalized counts. It can help us visualize the genes that are being differentially expressed. The blue points are the genes that are significantly differentially expressed. 

The direction of the triangle shows the direction of the logfold change. Genes in the upper right and lower right corners would have high mean of normalized counts as well as extreme (either higher or lower) logfold changes and would be interesting genes to follow up on.  

```{r}
# default 0.1 padj-based results
plotMA(results)
# 0.05 padj-based results 
plotMA(results0.05)
```







# Seperate this into a new .qmd file !!!!!!!!!!


# Weighted Gene Co-expression Network Analysis (WGCNA)

A subsequent reasonable step in our analysis is to run a WGCNA. When we run this on our significantly differentially expressed genes, it will create clusters/modules of similarly expressed (or co-expressed) genes. This can be helpful as it will direct us into interesting genes to look at for additional analysis like gene ontology classification, instead of starting with our thousands of inital genes. 

What we need to do here: 

* Construct a correlation network, calculate soft thresholds, and group genes into modules.
* Correlate with Traits: Identify which gene modules correlate with your conditions or phenotypes.

```{r}

```







