---
title: "OysterRNA24 DESeq2 Analysis"
author: "Callie Hundley"
format: 
  html:
    df-print: paged 
    code-fold: true 
    code-tools: true
editor: visual
---

# Project Overview 


## Loading packages 

```{r}
library(tidyverse)
library(lubridate)
library(readxl)
# BiocManager::install("DESeq2") # <- install only once!
library(DESeq2)
library(ggplot2)
```


## Loading data 

```{r}
# Read in the gene count data 
counts_data <- read_tsv("/work/clh162/OysterRNA24/hisat2_align/featureCounts_counts/gene_counts_matrix.txt") %>% 
  column_to_rownames(var = "Geneid") #assigning the gene names as the row names 
head(counts_data)


# Read in the sample data (defining sequenced samples)
coldata <- read_excel("/work/clh162/OysterRNA24/xRNA for BC Summer 2024.xlsx", 
    skip = 2) %>% #skipping first two rows of .xls file
  column_to_rownames(var = "Sample_ID") %>% #assigning the sample names as the row names 
  mutate(Month_collected = month(Date_Collected, label = TRUE, abbr = FALSE)) # creating streamlined column for graphing/analysis by date
head(coldata)
```

## DESeq2 Workflow 

### 1. Initial data tidying 

We are going to combine these two data sets off of the information that is shared  between them: the **names** of our sequenced samples. Because of this, we need to make sure that the names of the samples are identical between these two data frames.

From looking at our two data frames uploaded in the last code block, the names of the sample in the `counts_data` have an additional suffix on the end of each. We should shorten these to match the names of the samples in the `coldata` frame. 


```{r}
# rename the columns in the counts_data data frame to match sample names in the coldata data frame 
colnames(counts_data) <- sub("(_Gill).*", "\\1", colnames(counts_data))

colnames(counts_data)
rownames(coldata)

# make sure the row names in coldata matches column names in counts_data
all(rownames(coldata) %in% colnames(counts_data)) #TRUE

# make sure these names are also in the same order 
all(rownames(coldata) == colnames(counts_data)) #FALSE
match(colnames(counts_data), rownames(coldata)) 

# re-order 
counts_data[ ,rownames(coldata)]#this line works, so let's save it now
counts_data <- counts_data[ ,rownames(coldata)]
# Alternative way to write to match on the order of the counts_data column sample names
# coldata[colnames(counts_data), ] 
# coldata <- coldata[colnames(counts_data), ]

# re-check that the names are in the same order
all(rownames(coldata) == colnames(counts_data)) #TRUE

```

Now that we have matching column/row names and they are in the right order, we can construct our SummarizedExperiment (se) and DESeqDataSet (dds) objects! 

### 2. Create a se object 

**SummarizedExperiment (se) object:**
The idea of a se object is that it compiles all the gene expression data, gene features/annotations, and sample metadata into one cross-talking format. [The Bioconductor tutorial](https://carpentries-incubator.github.io/bioc-intro/60-next-steps.html#next-steps) talks about the value of creating a se object for inital omics data exploration. This tutorial summarizes how the se object is set up: 

* One (or more) assay(s) containing the quantitative omics data (expression data), stored as a matrix-like object. Features (genes, transcripts, proteins, …) are defined along the rows, and samples along the columns.
  * The count object(s) will be saved in assays slot

* A sample metadata slot containing sample information, stored as a data frame. Rows from this table represent samples (rows should be exact matches to the columns of the expression data which also hold the sample IDs). 
  * This sample information will be stored in `colData` slot (sample metadata)

* A feature metadata slot containing information describing each gene, stored as a data frame. The rows of this data frame match exactly the rows of the expression data (i.e. the gene names).
  * The rowranges object describing the genes will be stored in rowRanges slot (features metadata)


```{r}
se <- SummarizedExperiment(
    assays = list(counts = as.matrix(counts_data)),
    rowRanges = as(rowranges, "GRanges"),
    colData = coldata
)
```


The same Bioconductor tutorial referenced above eventually creates a **DESeq2** data object from the **se** object, however, you can also create a **DESeq2** object directly from a gene count matrix. 

### 2. Create a DESeq2 dataset object 

We will use the `DESeqDataSetFromMatrix()` command to create a dataset from our count data and sample data that is organized into the format required for the `DESeq()` command.  

```{r}
dds <- DESeqDataSetFromMatrix(countData = counts_data,
                       colData = coldata,
                       design = ~ Location) 
dds
```


### 3. Important pre-DESeq steps 

#### Removing unexpresed genes 

People will use different thresholds for this step, but let's remove genes that have fewer than 5 reads (this seems to be a very minimal threshold). 

```{r}
nrow(dds)

# Remove genes/rows that do not have > 5 total counts 
dds <- dds[rowSums(assay(dds, "counts")) > 5, ]
nrow(dds)
```




### 4. Run DESeq on dataset object 

```{r}
dds <- DESeq(dds)
```

Analyzing results 

```{r}
results <- results(dds, contrast = c("condition", "S", "C"))
```




# Seperate this into a new .qmd file !!!!!!!!!!


# Weighted Gene Co-expression Network Analysis (WGCNA)

A subsequent reasonable step in our analysis is to run a WGCNA. When we run this on our significantly differentially expressed genes, it will create clusters/modules of similarly expressed (or co-expressed) genes. This can be helpful as it will direct us into interesting genes to look at for additional analysis like gene ontology classification, instead of starting with our thousands of inital genes. 

What we need to do here: 

* Construct a correlation network, calculate soft thresholds, and group genes into modules.
* Correlate with Traits: Identify which gene modules correlate with your conditions or phenotypes.

```{r}

```







