---
title: "OysterRNA24 DESeq2 Analysis"
author: "Callie Hundley"
format: 
  html:
    df-print: paged 
    code-fold: true 
    code-tools: true
editor: visual
---

# Project Overview 


## Loading packages 

```{r}
library(tidyverse)
library(readxl)
# BiocManager::install("DESeq2") # <- install only once!
library(DESeq2)
library(ggplot2)
```


## Loading data 

```{r}
# Read in the gene count data 
counts_data <- read_tsv("/work/clh162/OysterRNA24/hisat2_align/featureCounts_counts/gene_counts_matrix.txt") %>% 
  column_to_rownames(var = "Geneid")
head(counts_data)

# Read in the sample data (defining sequenced samples)
coldata <- read_excel("/work/clh162/OysterRNA24/xRNA for BC Summer 2024.xlsx", 
    skip = 2) %>% #skipping first two rows of .xls file
  column_to_rownames(var = "Sample_ID") #assigning the sample names as the row names 

```

## DESeq2 Workflow 

### 1. Initial data tidying 

We are going to combine these two data sets off of the information that is shared  between them: the **names** of our sequenced samples. Because of this, we need to make sure that the names of the samples are identical between these two data frames.

From looking at our two data frames uploaded in the last code block, the names of the sample in the `counts_data` have an additional suffix on the end of each. We should shorten these to match the names of the samples in the `coldata` frame. 


```{r}
# rename the columns in the counts_data data frame to match sample names in the coldata data frame 
colnames(counts_data) <- sub("(_Gill).*", "\\1", colnames(counts_data))

colnames(counts_data)
rownames(coldata)

# make sure the row names in coldata matches column names in counts_data
all(rownames(coldata) %in% colnames(counts_data)) #TRUE

# make sure these names are also in the same order 
all(rownames(coldata) == colnames(counts_data)) #FALSE

# re-order 
coldata[colnames(counts_data), ] #this line works, so let's save it now
coldata <- coldata[colnames(counts_data), ]

# re-check that the names are in the same order
all(rownames(coldata) == colnames(counts_data)) #TRUE
```

Now that we have matching column/row names and they are in the right order, we can construct our DESeqDataSet object!


### 2. Create a DESeq2 dataset object 

We will use the `DESeqDataSetFromMatrix()` command to create a dataset from our count data and sample data that is organized into the format required for the `DESeq()` command.  

```{r}
dds <- DESeqDataSetFromMatrix(countData = counts_data,
                       colData = coldata,
                       design = ~ Location)
dds
```


### 3. Important pre-DESeq steps 





### 4. Run DESeq on dataset object 











