---
title: "OysterRNA24 Environmental Regression Analysis"
author: "Callie Hundley"
format: 
  html:
    df-print: paged 
    code-fold: true 
    code-tools: true
editor: visual
---

# Environmental Regression Analysis

### Purpose:
Test whether environmental conditions are associated with WGCNA module eigengenes and selected gene-expression patterns.

Analysis order:
1. Prepare sample-level environmental data
2. Explore environmental variables
3. Run module-level linear regressions
4. Identify the strongest module-environment associations
5. Examine selected genes using VST-normalized expression

### Load Packages

```{r}
library(readr)
library(readxl)
library(tidyverse)
library(lubridate)
#library(stringr)
#library(purrr)
```

## 1. Prepare sample-level environmental data

We will use a 7-day (week) window leading up to the date of collection of the tissue samples as our reference for our environmental data for a more accurate representation of environmental conditions at time of sampling. This seven day window will be the date of collection and the six days preceding it. 

### Load 2024 environmental cleaned data 

Here, we are loading the cleaned excel files containing our environmental data across the sites where our data sensors were placed for the BASS Connections project. Most of these files (salinity, pH, and DO) contain a column with the site that the logger data is associated with. However, for the precipitation data, this file isn't specific to just one site; it is for a general area that includes both of the sites being compared in this specific analysis.  

We are also standardizing our date column to follow the format of Date (without time) so they all match using the `as.Date()` function. Then we can get an average measurement of a given parameter for each day and use that to get our 7-day average. 

```{r}
# salinity 
sal_2024 <- read_excel("/work/clh162/Data+/Callie/bcdatafork/Salinity_2024.xlsx") %>% 
  dplyr::rename(Date = DateTime) %>%
  dplyr::mutate(Date = as.Date(Date)) %>% 
  dplyr::mutate(Farm = recode(Farm, "DUML" = "DAF", "CMAST" = "CMT"))  # to match our site names in coldata set
  
# dissolved oxygen 
do_2024 <- read_excel("/work/clh162/Data+/Callie/bcdatafork/DOsensor_2024.xlsx") %>% 
  dplyr::rename(Date = DateTime) %>%
  dplyr::mutate(Date = as.Date(Date)) %>% 
  dplyr::mutate(Farm = recode(Farm, "DUML" = "DAF", "CMAST" = "CMT"))

# pH
pH_2024 <- read_excel("/work/clh162/Data+/Callie/bcdatafork/pH_2024.xlsx") %>% 
  dplyr::rename(Date = DateTime) %>%
  dplyr::mutate(Date = as.Date(Date)) %>% 
  dplyr::mutate(Farm = recode(Farm, "DUML" = "DAF", "CMAST" = "CMT"))

# precipitation
precip_2024 <- read_excel("/work/clh162/Data+/Callie/bcdatafork/Precip_2024.xlsx") %>% 
  dplyr::mutate(
    Date = as.Date(DateTime),
    PrecipAccum = readr::parse_number(as.character(PrecipAccum)) # make this column read as numeric and get rid of any units written in with the values in the column
  )

# Temperature 
  # Was logged alongside each of the other three sensors. From looking at them, the temperatures seem fairly similar between the loggers. We can get an average temperature value for each of these sites from one of these sets (salinity, DO, and pH) or from an average of all the sets. 
  # 8/11/2026 : Decided to just pull mean temperature from each of these three loggers and then find the mean from all of them at each date (pool them together) -- to avoid having holes associated with one logger! 

```

Finally, if you don't already have your column data (sample metadata) loaded from earlier in the pipeline, you can reload it here as well -- we need this dataframe to identify which dates are the collection date for each tissue sample. 

```{r}
# coldata <- read_excel(
      "/work/clh162/OysterRNA24/xRNA for BC Summer 2024.xlsx",
#           col_types = c("text", "text", "text", "text", "text", 
#                          "date", "date", "numeric", 
#                         "numeric", "numeric", "numeric"),
#                         skip = 2) %>% #skipping first two rows of .xls file
#             column_to_rownames(var = "Sample_ID") %>% #assigning the sample names as the row names 
#             mutate(month_collected = month(Date_Collected, 
#                                            label = TRUE, 
#                                            abbr = FALSE)) # creating streamlined column for graphing/analysis by date
```

With the data loaded we can pull out our timeframes (around the collection dates) that we want to fill our data in for:

```{r}
# Pull out the distinct collection dates from sample metadata
collection_events <- coldata %>%
  distinct(
    Location,
    Date_Collected,
    month_collected
  ) %>% 
  dplyr::rename(Farm = Location) %>% # so that can communicate with our daily summaries
  arrange(Date_Collected) 

# Create one row for every required calendar day
# Collection day and the six preceding days = 7 total days
event_windows <- collection_events %>%
  mutate(
    window_start = Date_Collected - lubridate::days(6),
    window_end = Date_Collected
  ) %>%
  rowwise() %>%
  mutate(
    Date = list(
      seq(
        from = window_start,
        to = window_end,
        by = "day"
      )
    )
  ) %>%
  ungroup() %>%
  tidyr::unnest(Date)
```


### Create functions to return means of environmental parameters for each day 

Creating a "safe mean" -- using available values (bypassing NA values) when present and only returning `NA` for a day if there was absolutely no data logged for that variable. 

```{r}
safe_mean <- function(x) {
  if (all(is.na(x))) {
    return(NA_real_)
  }

  mean(x, na.rm = TRUE)
}
```

Additionally, for our precipitation values, we will want to be able to calculate total precipitation without running into an issue of `NA` values --> so we can create a "safe max" that bypasses NA values, because our precipitation data is given in cumulative values for each day. 

```{r}
safe_max <- function(x) {
  if (all(is.na(x))) {
    return(NA_real_)
  }

  max(x, na.rm = TRUE)
}
```


### Salinity / conductivity 

Goal: calculate 7-day average of salinity for the 7 days before the collection dates, at each site.

```{r}
salinity_daily <- sal_2024 %>%
  filter(
    Farm %in% c("DAF", "CMT"),
    !is.na(Date)
  ) %>%
  group_by(Farm, Date) %>%
  summarise(
    temperature_daily_mean_sal = safe_mean(Temp_C),
    conductivity_daily_mean = safe_mean(ConduScm),
    salinity_daily_mean = safe_mean(Sal_ppt),

    #temperature_observations = sum(!is.na(Temp_C)),
    #conductivity_observations = sum(!is.na(ConduScm)),
    #salinity_observations = sum(!is.na(Sal_ppt)),

    .groups = "drop"
  ) %>%
  arrange(Farm, Date)

# checking summary of salinity daily data
salinity_daily %>%
  group_by(Farm) %>%
  summarise(
    first_date = min(Date),
    last_date = max(Date),
    days_available = n_distinct(Date),
    .groups = "drop"
  )

# add the average values to summarized environmental table
event_window_data <- event_windows %>% 
  left_join(
  salinity_daily, 
  by = c("Farm", "Date")
  )

```



### pH

Goal: calculate 7-day average of pH for the 7 days before the collection dates, at each site.

```{r}

ph_daily <- pH_2024 %>%
  filter(
    Farm %in% c("DAF", "CMT"),
    !is.na(Date)
  ) %>%
  group_by(Farm, Date) %>%
  summarise(
    pH_daily_mean = safe_mean(pH),
    temperature_daily_mean_ph = safe_mean(Temp_C),
    # pH_observations = sum(!is.na(pH)),
    .groups = "drop"
  ) %>%
  arrange(Farm, Date)

# checking summary of pH daily data 
ph_daily %>%
  group_by(Farm) %>%
  summarise(
    first_date = min(Date),
    last_date = max(Date),
    days_available = n_distinct(Date),
    .groups = "drop"
  )


# add the average values to summarized environmental table 
event_window_data <- event_window_data %>%
  left_join(
  ph_daily,
  by = c("Farm", "Date"))
```


### Dissolved oxygen 

Goal: calculate 7-day average of DO for the 7 days before the collection dates, at each site.

```{r}
do_daily <- do_2024 %>%
  filter(
    Farm %in% c("DAF", "CMT"),
    !is.na(Date)
  ) %>%
  group_by(Farm, Date) %>%
  summarise(
    do_daily_mean = safe_mean(DomgL),
    temperature_daily_mean_do = safe_mean(Temp_C),
    # DO_observations = sum(!is.na(DomgL)),
    .groups = "drop"
  ) %>%
  arrange(Farm, Date)

# checking summary of DO daily data 
do_daily %>%
  group_by(Farm) %>%
  summarise(
    first_date = min(Date),
    last_date = max(Date),
    days_available = n_distinct(Date),
    .groups = "drop"
  )

# add the average values to summarized environmental table 
event_window_data <- event_window_data %>%
  left_join(
  do_daily,
  by = c("Farm", "Date"))
```


### Precipitation 

```{r}
# check that precipitation data is cumulative + RESETS between each day 
precip_2024 %>%
  filter(!is.na(Date)) %>%
  group_by(Date) %>%
  summarise(
    max_precip_accum = max(PrecipAccum, na.rm = TRUE),
    # observations = sum(!is.na(PrecipAccum)),
    .groups = "drop"
  ) %>%
  arrange(Date) %>%
  head(40)
  # It is reseting each day! 

precip_daily <- precip_2024 %>%
  filter(!is.na(Date)) %>%
  group_by(Date) %>%
  summarise(
    precipitation_daily_mm = safe_max(PrecipAccum),
    #precipitation_observations = sum(!is.na(PrecipAccum)),
    .groups = "drop"
  ) %>%
  arrange(Date)

# checking summary of precipitation daily data 
precip_daily %>%
  summarise(
    first_date = min(Date),
    last_date = max(Date),
    days_available = n_distinct(Date)
  )

# add the average values to summarized environmental table 
event_window_data <- event_window_data %>%
  left_join(
  precip_daily,
  by = c("Date")) # same precip assigned to both sites (not site-specific data)
```

### Temperature 

Now that we have mean values of temperature also pulled for each of our event window days from our three loggers, we can take the average of them and use that as our reference column for `temperature_daily_mean` in our `event_window_data`. 

```{r}
event_window_data <- event_window_data %>% 
  dplyr::mutate(
    temperature_daily_mean = rowMeans(
      pick(
        temperature_daily_mean_sal,
        temperature_daily_mean_ph,
        temperature_daily_mean_do
      ),
      na.rm = TRUE
    ),
    temperature_daily_mean = ifelse(
      is.nan(temperature_daily_mean),
      NA_real_,
      temperature_daily_mean
    )
  )
```

#### Save event window environmental data 

```{r}
write_csv(event_window_data,
          file = "collection_event_window_env_data.csv",
          path = "/work/clh162/OysterRNA24/DEG_analysis/WGCNA/collection_event_window_env_data.csv")

# re-load event window environmental data 
event_window_data <- read_csv("/work/clh162/OysterRNA24/DEG_analysis/WGCNA/collection_event_window_env_data.csv")
```


## **Investigating any gaps in environmental data during collection period** 

A reminder that for downstream analysis, we want 7-day windows of data to calculate mean values of environmental parameters associated with our six tissue collection dates (and gene expression of associated with these dates/samples).

We should aim to only fill gaps when we absolutely need to for our analysis (i.e. when we have NO data in the collection window). For our 7-day window, it would be better to have less data points covering the 7 days (need ideally 3) and use those to get the "7-day average" for environmental traits to run in our downstream analysis than to interpolate across the entire range and use (up to all 7 interpolated points) to calculate the average (because this would be **artificially shrinking our standard deviation**). Therefore, we will only use interpolation to fill in gaps where we have **NO** coverage for an environmental parameter in the whole 7-day collection window.

We can use an interpolation to do this. The `zoo` package in R (which stands for "Z's ordered observations") can be used to fill gaps in timeseries data with interpolation (e.g. linear, spline). We load that here first, and then begin this process. 

```{r}
# Install zoo package only if zoo is not already available
#if (!requireNamespace("zoo", quietly = TRUE)) {
#  install.packages("zoo")
#}

# Load zoo 
library(zoo)
```

Checking for gaps in environmental parameters during 7-day timeframe of data collection 

```{r}
missing_window_values <- event_window_data %>%
  select(
    Farm,
    Date_Collected,
    month_collected,
    Date,
    temperature_daily_mean, 
    conductivity_daily_mean,
    salinity_daily_mean,
    pH_daily_mean,
    do_daily_mean,
    precipitation_daily_mm
  ) %>%
  pivot_longer(
    cols = c(
      temperature_daily_mean,
      conductivity_daily_mean,
      salinity_daily_mean,
      pH_daily_mean,
      do_daily_mean,
      precipitation_daily_mm
    ),
    names_to = "environmental_variable",
    values_to = "value"
  ) %>%
  filter(is.na(value)) %>%
  select(
    Farm,
    Date_Collected,
    month_collected,
    missing_date = Date,
    environmental_variable
  ) %>%
  arrange(
    Date_Collected,
    environmental_variable,
    missing_date
  )

# quick peak at table of all missing values 
missing_window_values

```

From here, we can see that most of the 7-day spans have missing value gaps that are ~3-4 days long. Here, I am prioritizing having 3 points (days) with data in that 7-day span to have a baseline standard sample size to estimate the mean. Therefore, we should have two steps moving forward:

(1) let's interpolate (linearly -- because spline might infer excess variation that may not have happened) only for the groups that have 2 or fewer days adding to their calculated mean value. We can use the `na.approx()` function from the `zoo` package to run a linear interpolation.  

(2) Calculating a mean value in the 7 days leading up to sample collection using the **observed** values (and noting how many values are available/used to calculate this mean) --> can call this the **available-observation 7-day mean** (`mean_parameter_7d_observed`): Mean of observed pH measurements occurring within the seven days preceding and including tissue collection.


## Interpolation to fill large pH gap

The only instance of a gap that spans all 7 days leading up to sample collection is for the pH data at the CMAST (CMT) site with the August sample. 

```{r}
missing_window_values %>% 
  filter(Farm == "CMT",
         Date_Collected == "2024-08-19")
```

First, let's start by looking at the trends of our pH data (over the course of the entirety of the logged data we do have for the CMT site). This can help give us a good idea of the type of model to use to interpolate the data in this gap.  

#### Determining best-approach

```{r}
library(ggplot2)
library(dplyr)

cmt_ph <- ph_daily %>%
  filter(
    Farm == "CMT",
    Date >= as.Date("2024-06-07"),
    Date <= as.Date("2024-10-05")
  )

ggplot(cmt_ph, aes(x = Date, y = pH_daily_mean)) +
  geom_point() +
  geom_line() +
  
  # add lines designating the missing data ranges 
  geom_vline(
    xintercept = as.numeric(as.Date("2024-08-08")),
    color = "red"
  ) +
  geom_vline(
    xintercept = as.numeric(as.Date("2024-08-24")),
    color = "red"
  ) +
  
  # add lines designating the 7-day collection window
  geom_vline(
    xintercept = as.numeric(as.Date("2024-08-13")),
    linetype = "dashed",
    color = "blue"
  ) +
  geom_vline(
    xintercept = as.numeric(as.Date("2024-08-19")),
    linetype = "dashed",
    color = "blue"
  ) +
  theme_classic() +
  labs(
    title = "CMT pH: June–October 2024",
    x = "Date",
    y = "Daily mean pH"
  )
```

Let's make one more zoomed in plot 

```{r}
ggplot(cmt_ph, aes(x = Date, y = pH_daily_mean)) +
  geom_point() +
  geom_line() +
  coord_cartesian(
    xlim = as.Date(c("2024-07-25", "2024-09-10"))
  ) +
  geom_vline(
    xintercept = as.numeric(
      as.Date(c("2024-08-13", "2024-08-19"))
    ),
    linetype = "dashed",
    color = "blue"
  ) +
  geom_vline(
    xintercept = as.numeric(
      as.Date(c("2024-08-08", "2024-08-24"))
    ),
    color = "red"
  ) +
  theme_classic()
```

Looking at this graph, we can see that pH was approximately 8.0 immediately before the missing period (red lines) and approximately 8.0 immediately after it.

A **spline interpolation** would be trying to invent the shape of the pH trajectory between Aug 8 and Aug 24. But this data doesn't give us evidence that there was curvature during that period.

In fact, the most defensible inference from what we actually know is that the pH was very similar before and after this missing period. So using a **linear interpolation** would give us a very modest assumption of the data here and is defensibly a good one to use.

Because this is such a large gap in our data, it might be better to use groups of days preceding and following the gap rather than the single days on either end. Let's investigate the mean and sd of ranges of days before and after these timepoints to determine which grouping would be best to use.

We can write a function `get_summary` to do this!

```{r}
get_summary <- function(n) {
  
  before <- cmt_ph %>%
    filter(Date <= as.Date("2024-08-08")) %>%
    slice_tail(n = n) # get the last n days before the 8th 
  
  after <- cmt_ph %>%
    filter(Date >= as.Date("2024-08-24")) %>%
    slice_head(n = n) # get the first n days after the 24th 
  
  tibble(
    n_days = n,
    before_mean = mean(before$pH_daily_mean),
    before_sd = sd(before$pH_daily_mean),
    after_mean = mean(after$pH_daily_mean),
    after_sd = sd(after$pH_daily_mean),
    difference = mean(after$pH_daily_mean) -
                 mean(before$pH_daily_mean)
  )
}

bind_rows(
  get_summary(3),
  get_summary(5),
  get_summary(7)
)
```


We can see that these three sized groups of days results in very similar mean and standard deviation values. But the graph suggests that much of that lower "after" mean is coming from the decline occurring after August 24, rather than necessarily indicating that pH was already lower during August 13–19. So, because...

(1) there is not much of a different in mean/sd of pH values between the different sized groups of days before/after gap 

(2) we want to make as-few-assumptions as possible

and (3) want to limit the visible decline seen post-August-24 from influencing backward into your RNA exposure window (given that this happened post tissue sample collection)

the conceptual and defensible model to use in this case can be a **linear interpolation** using the **singular** dates pre and post this collection window.

Now, let's run the linear interpolation (since there is no mingap argument for the `na.approx()` function, it will compute the interpolation for all the missing values, but later in our calculation of 7d means, we can control when to use/ignore the interpolation values to make sure they are ONLY used for this large gap). 

#### Linear interpolation 

```{r}
# execute the linear interpolation 
ph_daily_interpolated <- ph_daily %>%
  group_by(Farm) %>%
  tidyr::complete(
    Date = seq(
      min(Date, na.rm = TRUE),
      max(Date, na.rm = TRUE),
      by = "day"
    )
  ) %>%
  arrange(Farm, Date) %>%
  mutate(
    
    # Was there an observed pH measurement?
    pH_observed = !is.na(pH_daily_mean),

    # Keep observed values; interpolate internal missing values
    pH_daily_filled = zoo::na.approx(
      pH_daily_mean,
      x = as.numeric(Date),
      na.rm = FALSE
    ),

    # Identify values that were generated by interpolation
    pH_was_interpolated =
      is.na(pH_daily_mean) &
      !is.na(pH_daily_filled)
  ) %>%
  ungroup()

# create subset of columns from ph_daily_interpolated to add to event_window_data (so not repeating columns shared)
ph_to_add <- ph_daily_interpolated %>% 
  select(
    Farm, #use as matching key
    Date, #use as matching key 
    pH_observed,
    pH_daily_filled,
    pH_was_interpolated
  )

# add to `event_window_data`
event_window_data_filled <- event_window_data %>%
  left_join(
    ph_to_add,
    by = c("Farm", "Date")
  )

# save this event_window data with interpolated pH added 
write_csv(event_window_data_filled,
          file = "collection_event_window_env_data_ph_interp.csv",
          path = "/work/clh162/OysterRNA24/DEG_analysis/WGCNA/collection_event_window_env_data_ph_interp.csv")

# load event_window data with interpolated pH added
event_window_data_filled <- read_csv("/work/clh162/OysterRNA24/DEG_analysis/WGCNA/collection_event_window_env_data_ph_interp.csv")
```


## Calculating 7-day means w/observed and interpolated data 


```{r}
summary_env_traits <- event_window_data_filled %>%
  group_by(
    Farm,
    Date_Collected,
    month_collected,
    window_start,
    window_end
  ) %>%
  summarise(
    
    # ============================================================
    # TEMPERATURE
    # ============================================================
    
    temperature_7d_observed_mean = mean(
      temperature_daily_mean,
      na.rm = TRUE
    ),
    
    temperature_n_observed = sum(
      !is.na(temperature_daily_mean)
    ),
    
    
    # ============================================================
    # SALINITY
    # ============================================================
    
    salinity_7d_observed_mean = mean(
      salinity_daily_mean,
      na.rm = TRUE
    ),
    
    salinity_n_observed = sum(
      !is.na(salinity_daily_mean)
    ),
    
    
    # ============================================================
    # CONDUCTIVITY
    # ============================================================
    
    conductivity_7d_observed_mean = mean(
      conductivity_daily_mean,
      na.rm = TRUE
    ),
    
    conductivity_n_observed = sum(
      !is.na(conductivity_daily_mean)
    ),
    
    
    # ============================================================
    # DISSOLVED OXYGEN
    # ============================================================
    
    do_7d_observed_mean = mean(
      do_daily_mean,
      na.rm = TRUE
    ),
    
    do_n_observed = sum(
      !is.na(do_daily_mean)
    ),
    
    
    # ============================================================
    # PRECIPITATION
    # ============================================================
    
    precipitation_7d_observed_mean = mean(
      precipitation_daily_mm,
      na.rm = TRUE
    ),
    
    precipitation_n_observed = sum(
      !is.na(precipitation_daily_mm)
    ),
    
    
    # ============================================================
    # pH
    # ============================================================
    
    pH_n_observed = sum(
      !is.na(pH_daily_mean)
    ),
    
    # Mean based ONLY on observed pH values
    pH_7d_observed_mean = if_else(
      pH_n_observed > 2,
      mean(
        pH_daily_mean,
        na.rm = TRUE
      ),
      NA_real_
    ),
    
    pH_n_interpolated = sum(
      is.na(pH_daily_mean) &
        !is.na(pH_daily_filled)
    ),
    
    # Mean based ONLY on interpolated pH values
    pH_7d_interpolated_mean = if_else(
      pH_n_observed == 0 &
        pH_n_interpolated == 7,
      mean(
        pH_daily_filled,
        na.rm = TRUE
      ),
      NA_real_
    ),
    
    .groups = "drop"
  )

# save summarized 7d average environmental trait data for WGCNA analysis 
write_csv(summary_env_traits,
          file = "envir_traits_7day_summary_2024.csv",
          path = "/work/clh162/OysterRNA24/DEG_analysis/WGCNA/envir_traits_7day_summary_2024.csv")
```


 
 
 
 



# ------------- EXTRA CODE -------------------------

## Interpolation to fill small gaps (AS OF 8/13 NOT running this code - instead just runing one larger interpolation for a singular full 7 day gap w/CMT August collection date)


##### Salinity / conductivity 

Note: we have some gaps of missing data for salinity + conductivity at both sites which overlap with the seven day averages we are trying to calculate. 

@ CMT: missing from 2024-07-19 and 2024-07-22 (overlap collection date 2024-07-22)
@ DAF: missing from 2024-06-15 and 2024-06-17 (overlap collection date 2024-06-21) and from 2024-07-20 and 2024-07-23 (overlap collection date 2024-07-23)   

Therefore we need to do some interpolation over these dates. 

The largest continuous gap for salinity/conductivity is for **10 days**. However, we are going to set the max interpolation no greater than 3 (per-notes above).  

```{r}
salinity_daily_interpolated <- salinity_daily %>%
  group_by(Farm) %>%
  tidyr::complete(
    Date = seq(
      min(Date, na.rm = TRUE),
      max(Date, na.rm = TRUE),
      by = "day"
    )
  ) %>%
  arrange(Farm, Date) %>%
  mutate(
    # Record whether the original daily measurement existed
    salinity_observed = !is.na(salinity_daily_mean),
    conductivity_observed = !is.na(conductivity_daily_mean),

    # Interpolate only internal gaps of ≤3 consecutive missing days.
    # Longer continuous gaps remain NA.
    salinity_daily_filled = zoo::na.approx(
      salinity_daily_mean,
      x = as.numeric(Date),
      maxgap = 3,
      na.rm = FALSE
    ),

    conductivity_daily_filled = zoo::na.approx(
      conductivity_daily_mean,
      x = as.numeric(Date),
      maxgap = 3,
      na.rm = FALSE
    ),

    # Flag values generated by interpolation
    salinity_imputed =
      is.na(salinity_daily_mean) &
      !is.na(salinity_daily_filled),

    conductivity_imputed =
      is.na(conductivity_daily_mean) &
      !is.na(conductivity_daily_filled)
  ) %>%
  ungroup()

# create dataframe of the interpolated values
salinity_values_imputed <- salinity_daily_interpolated %>%
  filter(salinity_imputed | conductivity_imputed) %>%
  select(
    Farm,
    Date,
    salinity_daily_mean,
    salinity_daily_filled,
    salinity_imputed,
    conductivity_daily_mean,
    conductivity_daily_filled,
    conductivity_imputed
  ) %>%
  arrange(Farm, Date)

# save dataframe of interpolated values
write_csv(salinity_values_imputed,
          file = "sal_conduct_daily_interpolated_smallgaps.csv",
          path = "/work/clh162/OysterRNA24/DEG_analysis/WGCNA/sal_conduct_daily_interpolated_smallgaps.csv")
```


##### pH

Note: we have some gaps of missing pH data at the CMT site which overlap with the seven day averages we are trying to calculate. 

@ CMT: between 2024-07-21 and 2024-07-22 (overlap collection date 2024-07-22), and between 2024-08-13 and 2024-08-19 (overlap collection date 2024-08-19)

Therefore we need to do some interpolation over these dates. We can use a spline interpolation to do this.

The largest gap for pH is for **continuous gaps of 15 days**. However, we are going to set the max interpolation no greater than 3 (per-notes above).  

```{r}
ph_daily_interpolated <- ph_daily %>%
  group_by(Farm) %>%
  tidyr::complete(
    Date = seq(
      min(Date, na.rm = TRUE),
      max(Date, na.rm = TRUE),
      by = "day"
    )
  ) %>%
  arrange(Farm, Date) %>%
  mutate(
    # Record whether the original daily measurement existed
    pH_observed = !is.na(pH_daily_mean),
    
    # Interpolate only internal gaps of ≤3 consecutive missing days.
    # Longer continuous gaps remain NA.
    pH_daily_filled = zoo::na.approx(
      pH_daily_mean,
      x = as.numeric(Date),
      maxgap = 3,
      na.rm = FALSE
    ),
    
    # Flag values generated by interpolation
    pH_imputed =
      is.na(pH_daily_mean) &
      !is.na(pH_daily_filled)
  ) %>%
  ungroup()

# create dataframe of the interpolated values
ph_values_imputed <- ph_daily_interpolated %>%
  filter(pH_imputed) %>%
  select(
    Farm,
    Date,
    pH_daily_mean,
    pH_daily_filled,
    pH_imputed
  ) %>%
  arrange(Farm, Date)


```

There were 0 days that were successfully interpolated using the maxgap length of 3 (so no interpolated dataframe saved like for `sal_conduct_daily`).

##### Dissolved oxygen 

No missing data here so we are good to go w/DO! 

##### Precipitation

No missing data here so we are good to go w/precipitation! 

##### Temperature 

By pulling across values of temperature from three sensors, there were no gaps / missing data here so we are good to go w/temperature! 

#### Interpolation summary 

How many values were interpolated? 

```{r}
# Create a table to summarize interpolation work 
tibble::tibble(
  variable = c("salinity", "conductivity", "pH"),
  values_imputed = c(
    sum(salinity_daily_interpolated$salinity_imputed),
    sum(salinity_daily_interpolated$conductivity_imputed),
    sum(ph_daily_interpolated$pH_imputed)
  )
)
```

A total of 6 days were interpolated for salinity and conductivity data, while 0 days were interpolated for pH. 


#### Filling in the gaps 

We want to update our `event_window_data` (saved as `collection_event_window_env_data.csv`) to fill in the holes with our interpolated data covering short gaps. We can prepared our daily and daily_interpolated datasets to do this:

```{r}
# Prepare interpolated salinity and conductivity data
salinity_daily_filled <- salinity_daily_interpolated %>%
  transmute( 
    # the transmute function is like a combination of mutate() (creating new columns) and select() (automatically dropping any columns in the provided df that are NOT explicity mentioned)
    Farm = Farm, # new column = old column
    Date = Date,
  ## Salinity 
    # Assign salinity/conductivity values (including those that were interpolated)
    salinity_daily_mean = salinity_daily_filled,
    
    # Flag values generated by interpolation
    salinity_was_interpolated = salinity_imputed,
  
  ## Conductivity 
    # Assign salinity/conductivity values (including those that were interpolated)
    conductivity_daily_mean = conductivity_daily_filled,
    
    # Flag values generated by interpolation
    conductivity_was_interpolated = conductivity_imputed
  )

  # NO pH data was successfully linearly interpolated 
# Prepare interpolated pH data
# ph_daily_filled <- ph_daily_interpolated %>%
#   transmute(
#     Farm = Farm,
#     Date = Date,
# 
#     # Assign pH values (including those that were interpolated)
#     pH_daily_value = pH_daily_filled,
#     
#     # Flag values generated by interpolation
#     pH_was_imputed = pH_imputed
#   )
```

```{r}
# Create a new filled out event_window_data dataset 

# First, drop all the salinity and conductivity columns that will be replaced (the means) 
event_window_data_filled <- event_window_data %>% 
  select(!c(
    conductivity_daily_mean,
    salinity_daily_mean,
    pH_daily_mean)
  )

# Add in the prepared interpolated salinity, conductivity, and pH columns 
event_window_data_filled <- event_window_data_filled %>% 
  dplyr::left_join(
    salinity_daily_filled,
    by = c("Farm", "Date")
  ) 


# Save filled event window environmental data 
write_csv(event_window_data_filled,
          file = "collection_event_window_env_data_full.csv",
          path = "/work/clh162/OysterRNA24/DEG_analysis/WGCNA/collection_event_window_env_data_full.csv")
```


# Environmental Data correlations 

```{r}
# create table of correlation coefficients for environmental variables
correlation_table <- data.frame(
  Gene = names(temp_correlation),
  Temperature = temp_correlation,
  Salinity = salinity_correlation,
  DO = do_correlation,
  pH = ph_correlation,
  row.names = NULL
)
correlation_table
```

Now that we have our 7-day averages for our environmental parameters paired with our collection dates of tissue samples at each site, we can calculate correlation coefficients between our top DEGs and these environmental parameters. 

### Salinity

```{r}
# calculate correlation coefficients
salinity_correlation <- sapply(
  top_month_genes, # need to find equivalent of this in Sophie's code to my code 
  function(gene)
    cor(
      assay(vsd_time)[gene, ],
      env_data$Salinity_7d_Avg,
      method = "spearman"
    )
)
salinity_correlation

# create PCA plot for salinity and location
pcaData$Salinity <- env_data$Salinity_7d_Avg
pcaData$Location <- env_data$Location

ggplot(
  pcaData,
  aes(
    PC1,
    PC2,
    color = Salinity,
    shape = Location
  )
) +
  geom_point(size = 4) +
  theme_classic()
```

### pH 

```{r}
# calculate correlation coefficients
ph_correlation <- sapply(
  top_month_genes,
  function(gene)
    cor(
      assay(vsd_time)[gene, ],
      env_data$pH_7d_Avg,
      method = "spearman",
      use = "complete.obs"
    )
)
ph_correlation

# create PCA plot for ph and location
pcaData$pH <- env_data$pH_7d_Avg

ggplot(
  pcaData,
  aes(
    PC1,
    PC2,
    color = pH,
    shape = Location
  )
) +
  geom_point(size = 4) +
  theme_classic()
```

### Dissolved oxygen 

```{r}
# calculate correlation coefficients
do_correlation <- sapply(
  top_month_genes,
  function(gene)
    cor(
      assay(vsd_time)[gene, ],
      env_data$DO_7d_Avg,
      method = "spearman",
      use = "complete.obs"
    )
)
all(colnames(vsd_time) == env_data$Sample)
do_correlation

# create PCA plot for dissolved oxygen and location
pcaData$DO <- env_data$DO_7d_Avg
pcaData$Location <- env_data$Location

ggplot(
  pcaData,
  aes(
    PC1,
    PC2,
    color = DO,
    shape = Location
  )
) +
  geom_point(size = 4) +
  theme_classic()
```



## 2. Explore environmental variables



## 3. Run module-level linear regressions



## 4. Identify the strongest module-environment associations




## 5. Examine selected genes using VST-normalized expression


