---
title: "OysterRNA24 GO Analysis"
author: "Callie Hundley"
format: 
  html:
    df-print: paged 
    code-fold: true 
    code-tools: true
editor: visual
---


# Gene ontology analysis 

To conduct GO analysis we need a list of gene symbols, entrez ids, or ensembl ids. Depending on the type of information you have, you may need to convert this information into a different format. Gene annotation involves identifying information about the locations of genes along with their function, in addition to information about other genetic and noncoding features. 

Because we do not have **gene annotations** (i.e. gene ids) for this new *C. virginica* genome, we need to create them. We can use the aligner `DIAMOND` to do this. `DIAMOND` is great because it allows you to run actual nucleotide/protein sequences against a database to find alignment/matches for annotations instead of just using standard gene IDs. `DIAMOND` is not the same as `BLASTP`; it does runs similar searches and returns similar files as `BLASTP`, but does it at a much faster rate!  

So let's run a search with `DIAMOND` (in terminal -- the code below is just for reference). 

First we need to install diamond: 
^ You can do this by having the executable file of the program (diamond.exe) downloaded into the folder where your gene.fasta files and reference files are. 

```{r}
# installing diamond (within our RNA-seq environment)

# conda activate RNA-seq 
# conda install bioconda::diamond
```

What `BLASTP` does is that it spends time looks for a 3-sequential sequences somewhere in the file being queried that matches any reference database; this can end up taking a long time (up to hours!). `DIAMOND` bypasses this by having a function to compile an a index to sequence databases that allows for faster search. So next, we will make this index for each of our reference genomes; these will be saved as `.dmnd` files.

We will make this index for the new (2025) genome. 

```{r}
# make index once 
# generic format 
  # diamond makedb --in newgenomeproteome.fasta -d newgenomeproteomedb
# specific line (actually run - see 06_diamond.sh)
  # diamond makedb --in /work/clh162/OysterRNA24/hisat2_align/Cv_genome_RU_2025_shared/Cvi_RU25.pep -d Cvi_RU25proteome
```

Finally, using this created index, we can run a `DIAMOND` search. 

```{r}
# run search for each protein -- looking for every sequence in the index.dmnd file that has great similarity to the individual protein provided.  
# genetic format 
  # diamond blastp -q oldgenome.fa -db newgenomeproteomedb.dmnd -o old_to_new_diamond.tsv --evalue 1e-10 --max-target-seqs 1 --ultra-sensitive --outfmt 6 qseqid sseqid pident length mismatch gapopen qstart qend sstart send evalue bitscore
# specific line (actually run - see 06_diamond.sh)
  # diamond blastp -q /work/clh162/OysterRNA24/DEG_analysis/GO/Cvi_3.0_GCF_002022765.2/protein.faa --db /work/clh162/OysterRNA24/DEG_analysis/GO/Cvi_RU25proteome.dmnd -o old_to_new_diamond --evalue 1e-10 --max-target-seqs 1 --ultra-sensitive --outfmt 6 qseqid sseqid pident length mismatch gapopen qstart qend sstart send evalue bitscore
```

Looking at the output file, we will be primarily interested in a few reported results from the 12 tab output values: 

* Tab 3: Sequence identity -- The % of amino acids identical between the querey sequence and the match sequence found within the database.dmnd file.
* Tab 11: E score -- How likely are you to produce a hit of similar or better quality than the one produced here if you were to run this query search against random sequences. **Lower = better** 
* Tab 12: Bit score -- Think of as the # of points scored by a homology search. **Higher = better**

Q: What if `DIAMOND` returns more than one line for the `blastp` search?

* If `DIAMOND` doesn't return anything, it would mean that there is no ortholog for that gene 

* If `DIAMOND` returns more than one line, could be a sign of genome duplication, including the gene that is being queried against the database. 

Now that we have gene annotation files from the `DIAMOND` program, we can carry out our **GO analysis**. 


```{r}
# Installing packages -- RUN THIS CODE CHUNK ONCE 
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(version = "3.20") #newest version is 3.23 but need to use 3.20 with R version 4.4
```

```{r}
# Loading libraries 
library(clusterProfiler)
library(AnnotationDbi)
```

Let's focus on running our GO analysis only on a subset of significantly expressed genes. First to get the significantly differentially expressed genes, we can subset from our results dataframe from the dds. 

```{r}
sign_DEG <- na.omit(results)
sign_DEG <- sign_DEG[sign_DEG$padj < 0.05 & sign_DEG$baseMean > 50] # need to decide what threshold values we want to use for defining significance
```


## GO analysis for positively differentially expressed genes that have a value above 0.5 for log2FoldChange 

First, we can subset our significant DEG dataset (sign_DEG) to target this group. 

```{r}
sign_DEG[sign_DEG$log2FoldChange > 0.5, ]
```

Then we need to pull out a list of the gene names from this subset.

```{r}
rownames(sign_DEG[sign_DEG$log2FoldChange > 0.5, ])
# save this list 
list <- rownames(sign_DEG[sign_DEG$log2FoldChange > 0.5, ])
```

Using this list we can now run GO analysis

```{r}
GO_results <- clusterProfiler::enrichGO(genes = genes_to_test,
                                        OrgDb = ,
                                        keyType = )
```
