---
title: "OysterRNA24 GO Analysis"
author: "Callie Hundley"
format: 
  html:
    df-print: paged 
    code-fold: true 
    code-tools: true
editor: visual
---

# Initial Packages 

```{r}
library(readr)
library(dplyr)
```


# Gene ontology analysis 

To conduct GO analysis we need a list of gene symbols, entrez ids, or ensembl ids to figure out what GO terms are associated with each gene. Depending on the type of information you have, you may need to convert this information into a different format. Gene annotation involves identifying information about the locations of genes along with their function, in addition to information about other genetic and noncoding features. 

Because we do not have **gene annotations** (i.e. gene ids) for this new *C. virginica* genome, we need to find the "best matches" of named genes in the old genome. We can use the aligner `DIAMOND` to do this. `DIAMOND` is great because it allows you to run actual nucleotide/protein sequences against a database to find alignment/matches for annotations instead of just using standard gene IDs. `DIAMOND` is not the same as `BLASTP`; it does runs similar searches and returns similar files as `BLASTP`, but does it at a much faster rate!  

So let's run a search with `DIAMOND` (in terminal -- the code below is just for reference). 

First we need to install diamond: 
^ You can do this by having the executable file of the program (diamond.exe) downloaded into the folder where your gene.fasta files and reference files are. 

```{r}
# installing diamond (within our RNA-seq environment)

# conda activate RNA-seq 
# conda install bioconda::diamond
```

What `BLASTP` does is that it spends time looks for a 3-sequential (bp) sequences somewhere in the file being queried that matches any reference database; this can end up taking a long time (up to hours!). `DIAMOND` bypasses this by having a function to compile an a index to sequence databases that allows for faster search. So next, we will make this index for the old (already partially annotated) genome which we can then query our new genome against; these index files will be saved as `.dmnd` files.

We will make this index for the old (Yale) genome using the `diamond makedb` command and a protein FASTA file (`.faa` works) from the old genome. What this command is doing is extracting the proteins, which will serve as our index database -- a db that will serve as a searchable reference for the genes in our new genome.   

```{r}
# make index once 
# generic format 
  # diamond makedb --in old_genome_proteins.fasta -d old_genome_db
# specific line (actually run - see 06_diamond.sh)
  # diamond makedb --in /work/clh162/OysterRNA24/DEG_analysis/GO/Cvi_RU17_ncbi_dataset/data/GCF_002022765.2/protein.faa -d /work/clh162/OysterRNA24/DEG_analysis/GO/diamond/Cvi_RU17_proteome
```

Finally, using this created index, we can run a `diamond blastp` search. We are taking our new genome (protein file - not annotated) and create a list of predicted proteins (protein each one is most similar to) using the old genome protein database as a reference. The output of this command will give us a list that says something like "newgene1 is most similar to oldgene12" (with a list of metrics sharing how similar they are).

```{r}
# run search for each protein -- looking for every sequence in the index.dmnd file that has great similarity to the individual protein from our new genome provided.  
# genetic format 
  # diamond blastp -q new_genome_proteins.faa -db old_genome_db.dmnd -o old_to_new_diamond.tsv --evalue 1e-10 --max-target-seqs 1 --ultra-sensitive --outfmt 6 qseqid sseqid pident length mismatch gapopen qstart qend sstart send evalue bitscore

# specific line (actually run - see 06_diamond.sh)
  # diamond blastp -q /work/clh162/OysterRNA24/hisat2_align/Cv_genome_RU_2025_shared/Cvi_RU25.pep --db /work/clh162/OysterRNA24/DEG_analysis/GO/diamond/Cvi_RU17_proteome.dmnd -o /work/clh162/OysterRNA24/DEG_analysis/GO/diamond/RU17_to_RU25_diamond.tsv --evalue 1e-10 --max-target-seqs 1 --ultra-sensitive --outfmt 6 qseqid sseqid pident length mismatch gapopen qstart qend sstart send evalue bitscore
```

Looking at the output file, we will be primarily interested in a few reported results from the 12 tab output values: 

* Tab 3: Sequence identity -- The % of amino acids identical between the querey sequence and the match sequence found within the database.dmnd file.
* Tab 11: E score -- How likely are you to produce a hit of similar or better quality than the one produced here if you were to run this query search against random sequences. **Lower = better** 
* Tab 12: Bit score -- Think of as the # of points scored by a homology search. **Higher = better**

Q: What if `DIAMOND` returns more than one line for the `blastp` search?

* If `DIAMOND` doesn't return anything, it would mean that there is no ortholog for that gene 

* If `DIAMOND` returns more than one line, could be a sign of genome duplication, including the gene that is being queried against the database. However, we are controlling the number of matches returned in our code above to be just one gene from the old genome.

However, if we look closely, this `diamond` search sometimes returned the same gene in the old genome to match with multiple genes in the new genome. We can control for this by limiting only ONE line be returned for each of our new genes -- we can specify that we want this one line to be the one with the lowest E-value (tab 11). <- **Need to find some way to actually do this!**

```{r}
# how to do this....
  #6/30/26 skipped for now 

```


# Matching gene names 

^ The work done below was run at the command line in the stated working directory

The `RU17_to_RU25_diamond.tsv` file gives us an output that has our gene names (RU25 genome) and best-match (orthologs) proteins found in the old Rutgers genome (RU17). We have a file associated with the old genome (RU17) that contains these protein names and corresponding gene ID. So what we need to do is match our gene names to their best-match genes from the old genome by matching on the protein_ID that is present in both datasets. 

First, we need a clean file that just has the old genome gene IDs and protein IDs. 

```{r}
# Working directory: /work/clh162/OysterRNA24/DEG_analysis/GO/Cvi_RU17_ncbi_dataset/data/GCF_002022765.2

# create file to hold stripped gene ID and protein ID information from the cds_from_genomic.fna in the 2017 RU
  # touch Cvi_RU17_gene_proteinID.csv
  # echo gene,protein_id > Cvi_RU17_gene_proteinID.csv

# strip information and append to the .csv file
  #awk '/^>/ {
   # match($0, /\[gene=([^]]+)\]/, g)
   # match($0, /\[protein_id=([^]]+)\]/, p)
   # print g[1] "," p[1]
#}' cds_from_genomic.fna >> Cvi_RU17_gene_proteinID.csv
```

Now that that is saved, we can load it in here:

```{r}
Cvi_RU17_gene_protein <- read_csv("/work/clh162/OysterRNA24/DEG_analysis/GO/Cvi_RU17_ncbi_dataset/data/GCF_002022765.2/Cvi_RU17_gene_proteinID.csv")
```

Next we need to match these names using our diamond output file. This file was saved in `.tsv` format, and we can load it as a dataframe here!

```{r}
# Load diamond translation .tsv file 
column_names <- c("RU25_gene", "protein_id", "pident", "length", "mismatch", "gapopen", "qstart", "qend", "sstart", "send", "evalue", "bitscore")

RU17_to_RU25_diamond <- read_tsv("/work/clh162/OysterRNA24/DEG_analysis/GO/diamond/RU17_to_RU25_diamond.tsv", col_names = column_names)
```

Also, the names are slightly mis-matched from the gene names that would be found in our list of DEGs from the DESeq `results()` output. So we need to fix those names here as well. 

```{r}
library(stringr)

# replace the '.model.' with '.TU.' to match our results file  
RU17_to_RU25_diamond$RU25_gene <- str_replace(RU17_to_RU25_diamond$RU25_gene, "\\.model\\.", ".TU.")
```


We can join these two loaded data sets on the column of `protein_id` because they both have it.  

```{r}
# left join - pull evm gene ID over to the stripped Cvi_RU17_gene_protein dataset  
left_join(x = RU17_to_RU25_diamond, # will return all our genes that found a match in the diamond search 
          y = Cvi_RU17_gene_protein, # and their corresponding LOC gene ID names (ENSEMBL)
          by = "protein_id",
          )


# Subset of this joined dataset that just has the matching gene IDs
RU17_RU25_gene_IDs <- left_join(x = RU17_to_RU25_diamond, 
                             y = Cvi_RU17_gene_protein,
                             by = "protein_id",
                             ) %>% 
  select(RU25_gene, protein_id, gene)

# Save this subset dataset for analysis of DEG results data
write_csv(RU17_RU25_gene_IDs,
          file = "RU17_RU25_gene_IDs",
          path = "/work/clh162/OysterRNA24/DEG_analysis/GO/RU17_RU25_gene_IDs")
```

Now we have the list that helps us correspond the 2017 sequenced genome ENSEMBL gene names to the 2025 sequenced genome gene names. For each group of our DE analysis results that we want to run through GO, we can use this file to add the old ENSEMBL gene IDs into our `results()` output pre-GO analysis. With these ENSEMBL gene IDs assigned, we can pair corresponding GO terms to our DE genes!  


Now that we have homologous gene IDs and information about the gene annotations from the `DIAMOND` program and gene name matching, we can carry out our **GO analysis**. 

# GO analysis 

Like for the other steps in our pipeline, there are a lot of different packages that can be used to conduct GO analysis. Here, we are going to try using the `goseq()` command from the ____ package. 

First, let's load all the packages we need. 

```{r}
# Installing packages -- RUN THIS CODE CHUNK ONCE 
#if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
#BiocManager::install(version = "3.20") #newest version is 3.23 but need to use 3.20 with R version 4.4
#library(BiocManager, lib.loc = "/usr/local/lib/R/site-library")

# BiocManager::install("clusterProfiler")
# BiocManager::install("biomaRt")
# BiocManager::install("goseq")
```



```{r}
# Loading libraries 
install.packages("ggplot2", dependencies = TRUE)
library(clusterProfiler) #built-in functions to generate various visualizations (incl. bubble plots, dot plots, enrichment maps)
library(AnnotationDbi)
# Packages below installed with BiocManager
library(biomaRt)
library(goseq)
```

Was getting the following error when trying to load `biomaRt` so then ran the following code chunk to fix what appeared to be version mismatch of these packages: 

**Error: package or namespace load failed for ‘biomaRt’:
 object ‘filter_out’ is not exported by 'namespace:dplyr'**
 
```{r}
#remove.packages(c("dplyr", "dbplyr", "biomaRt"))
#install.packages("dplyr")
#BiocManager::install("biomaRt")
```


## DAF vs. CMT June

Let's load + prepare to run our results of differential gene expression between Duke Aquafarm (DAF) and CMAST farm (CMT) in June through the GO enrichment analysis pipeline. 

### Loading DESeq results + assigning old gene IDs 

First we will load the saved results `.csv` file and add the gene name translation file to it via a join.  

```{r}
# results from June by site
res_0.05_DAF_vs_CMT_June <- read_csv("/work/clh162/OysterRNA24/DEG_analysis/DESeq2/res_0.05_DAF_vs_CMT_June")
# check structure
head(res_0.05_DAF_vs_CMT_June)

# load old to new gene ID file for ENSEMBL ID communication
RU17_RU25_gene_IDs <- read_csv("/work/clh162/OysterRNA24/DEG_analysis/GO/RU17_RU25_gene_IDs")

# run a left_join() to get all the ENSEMBL gene names added to all the genes in this results set 
res_0.05_DAF_vs_CMT_June <- left_join(x = res_0.05_DAF_vs_CMT_June,
          y = RU17_RU25_gene_IDs,
          by = join_by(gene_id == RU25_gene)) # column names for RU25 genes are different between the two data sets 
res_0.05_DAF_vs_CMT_June

# how many of our DE genes don't have a homologous match (ENSEMBL ID) in the RU17 genome? 
res_0.05_DAF_vs_CMT_June %>% filter(is.na(gene)) # 2,123
# remove DE genes w/NA assignment in ENSEMBL_ID value 
#res_0.05_DAF_vs_CMT_June <- res_0.05_DAF_vs_CMT_June %>% filter(!is.na(gene))
  # brings us down to 19,203 DEGs with a "best match" to old genome 

```

Notice that there are a few (~2K) DEGs in this set that do not have a good match for a gene from the old Rutgers 2017 genome (have NAs in the gene column). 

Now that we have all the packages that we need loaded and our DE data, we can start our GO analysis. Next, we need to get the GO terms associated with the DE genes in this group of results. However here we run into another issue since we are using a brand-new sequenced genome: it doesn't have GO terms assigned yet! So we either need to create them (using a program like `eggnog()`) or use published GO terms associated with an older genome. We will go with the latter and pair our gene_ids with GO terms from the Rutgers 2017 genome. We can do this by establishing a connection to the Ensembl website/database where they are published.
 
### Connect GO terms using ENSEMBL gene IDs

```{r}
# Establish connection to Ensembl Genes and the Rutgers 2017 genome dataset
mart <- useMart("metazoa_mart", 
                dataset = "cvgca002022765v4_eg_gene", # old genome name 
                host = "https://metazoa.ensembl.org")

# Pull the Gene ontology (GO) IDs associated with each gene in Rutgers 2017 genome dataset 
attrs <- getBM(attributes = c("ensembl_gene_id", "transcript_length", "go_id", "namespace_1003"), 
               mart = mart)

# subset for list of the ensembl_gene_ids and go_ids for only the genes that have attributes listed (remove genes without go ids listed for them)
gene2cat <- attrs[attrs$go_id != "", c("ensembl_gene_id", "go_id")]

# get the length for each ensembl_gene_id (one row for each gene)
gene_lengths <- aggregate(transcript_length ~ ensembl_gene_id, data = attrs, max)

# how many ensembl_gene_ids genes in the gene_lengths dataset are represented in our DEG results?  
gene_lengths %>% 
  filter(ensembl_gene_id %in% res_0.05_DAF_vs_CMT_June$gene)
  # 39,530 genes originally pulled from ENSEMBL, only 16,345 of them match to our DEG results 

# put length information into vector format  
lengths_vector <- gene_lengths$transcript_length
# add ensembl_gene_id names to the lengths data 
names(lengths_vector) <- gene_lengths$ensembl_gene_id
```

Next, we need to put information about significance of our differentially expressed genes into binary format:

```{r}
# Create DE gene vector where 1 = significant and 0 = not
DE_genes_vector <- as.integer(
  res_0.05_DAF_vs_CMT_June$padj < 0.05 &
  !is.na(res_0.05_DAF_vs_CMT_June$padj) # exclude DE genes that do not have a padj value 
)

# add ensembl_gene_id names to the lengths data 
names(DE_genes_vector) <- res_0.05_DAF_vs_CMT_June$gene
```

In order to run `goseq()`, the gene length and differential expression vectors need to be the same length (i.e. one row in each vector for each gene). This step limits the rows in each vector to only include those genes that are present in both vectors. 

```{r}
# Now restrict vectors to overlap with this DESeq result universe:
# create a list of common_genes shared between these two vectors 
common_genes <- intersect(names(DE_genes_vector), names(lengths_vector))

DE_genes_vector <- DE_genes_vector[common_genes]
lengths_vector <- lengths_vector[common_genes]
  # cut down from 21326 in DE_genes_vector and 39530 in lengths_vector to 16345 (the final # we expected from above)

# Make sure our vector files are the same length: the vectors provided to the nullp() command need to have the same length 
length(DE_genes_vector)
length(lengths_vector)

# another check: all names are the same 
all(names(DE_genes_vector) == names(lengths_vector))

# Final sanity check 
stopifnot(length(DE_genes_vector) == length(lengths_vector))
stopifnot(all(names(DE_genes_vector) == names(lengths_vector)))

```

### Investigation of differential expression + significance 

Let's look at how many genes are differentially expressed. When Henry ran this he was using the shrinkage values way back during QC analysis of samples. 

```{r}
#How many genes are differentially expressed?
attach(as.data.frame(LFC))
#The total number of DEGs with an adjusted p-value<0.05
summary(LFC, alpha=0.05)
#The total number of DEGs with an adjusted p-value<0.05 AND absolute fold-change > 2
sum(!is.na(padj) & padj < 0.05 & abs(log2FoldChange) >2)

#Decreased expression:
sum(!is.na(padj) & padj < 0.05 & log2FoldChange <0) #any fold-change
sum(!is.na(padj) & padj < 0.05 & log2FoldChange <(-2)) #fold-change greater than 2
#Increased expression
sum(!is.na(padj) & padj < 0.05 & log2FoldChange >0) #any fold-change
sum(!is.na(padj) & padj < 0.05 & log2FoldChange >2) #fold-change greater than 2
```

### Running goseq()

Finally, we are prepared to run `goseq()`

```{r}
# account for the fact that longer genes are easier to sequence/detect
pwf <- nullp(DEgenes = DE_genes_vector, bias.data = lengths_vector)

# run goseq
GO.wall <- goseq(pwf = pwf, 
                 gene2cat = gene2cat,
                 test.cats = ,
                 use_genes_without_cat = FALSE)
  # 3730 genes w/out GO terms -- excluded from p-value calculation
  # 5501 GO terms (rows)

# apply multiple testing correction method of Benjamini-Hochberg (BH) to control the False Discovery Rate (FDR) (aka limit type II errors)
GO.wall$padj <- p.adjust(GO.wall$over_represented_pvalue, method = "BH")

# Filter for significant terms (e.g., padj < 0.05)
enriched_GO <- GO.wall[GO.wall$padj < 0.05, ] # zero GO terms are significantly abundant 

# save goseq() results to a .csv file 
write_csv(GO.wall,
          file = "GOres_0.05_DAF_vs_CMT_June",
          path = "/work/clh162/OysterRNA24/DEG_analysis/GO/goseq_results/GOres_0.05_DAF_vs_CMT_June")
```


## DAF vs. CMT July 

### Loading DESeq results + assigning old gene IDs 

```{r}
# results from July by site
res_0.05_DAF_vs_CMT_July <- read_csv("/work/clh162/OysterRNA24/DEG_analysis/DESeq2/res_0.05_DAF_vs_CMT_July")

# run a left_join() to get all the ENSEMBL gene names added to all the genes in this results set 
res_0.05_DAF_vs_CMT_July <- left_join(x = res_0.05_DAF_vs_CMT_July,
          y = RU17_RU25_gene_IDs,
          by = join_by(gene_id == RU25_gene)) # column names for RU25 genes are different between the two data sets 
res_0.05_DAF_vs_CMT_July

# how many of our DE genes don't have a homologous match (ENSEMBL ID) in the RU17 genome? 
res_0.05_DAF_vs_CMT_July %>% filter(is.na(gene)) # still 2,123 genes don't have homologous match (ENSEMBL ID)
```

### Connect GO terms using ENSEMBL gene IDs

```{r}
# how many ensembl_gene_ids genes in the gene_lengths dataset are represented in our DEG results?  
gene_lengths %>% 
  filter(ensembl_gene_id %in% res_0.05_DAF_vs_CMT_July$gene)
  # 39,530 genes originally pulled from ENSEMBL, still only 16,345 of them match to our DEG results 

# RESETING LENGTHS_VECTOR with the original gene_lengths file 
  # put length information into vector format  
lengths_vector <- gene_lengths$transcript_length
# add ensembl_gene_id names to the lengths data 
names(lengths_vector) <- gene_lengths$ensembl_gene_id

# Create DE gene vector where 1 = significant and 0 = not
DE_genes_vector <- as.integer(
  res_0.05_DAF_vs_CMT_July$padj < 0.05 &
  !is.na(res_0.05_DAF_vs_CMT_July$padj) # exclude DE genes that do not have a padj value 
)

# add ensembl_gene_id names to the lengths data 
names(DE_genes_vector) <- res_0.05_DAF_vs_CMT_July$gene



# Now restrict vectors to overlap with this DESeq result universe:
# create a list of common_genes shared between these two vectors 
common_genes <- intersect(names(DE_genes_vector), names(lengths_vector))

DE_genes_vector <- DE_genes_vector[common_genes]
lengths_vector <- lengths_vector[common_genes]
  # cut down from 21326 in DE_genes_vector and 39530 in lengths_vector to 16345 (the final # we expected from above)

# Make sure our vector files are the same length: the vectors provided to the nullp() command need to have the same length 
length(DE_genes_vector)
length(lengths_vector)

# another check: all names are the same 
all(names(DE_genes_vector) == names(lengths_vector))

# Final sanity check 
stopifnot(length(DE_genes_vector) == length(lengths_vector))
stopifnot(all(names(DE_genes_vector) == names(lengths_vector)))
```

### Investigation of differential expression + significance

```{r}

```

### Running goseq()

```{r}
# account for the fact that longer genes are easier to sequence/detect
pwf <- nullp(DEgenes = DE_genes_vector, bias.data = lengths_vector)

# run goseq
GO.wall <- goseq(pwf = pwf, 
                 gene2cat = gene2cat,
                 test.cats = ,
                 use_genes_without_cat = FALSE)
  # 3730 genes w/out GO terms -- excluded from p-value calculation
  # 5501 GO terms (rows)

# apply multiple testing correction method of Benjamini-Hochberg (BH) to control the False Discovery Rate (FDR) (aka limit type II errors)
GO.wall$padj <- p.adjust(GO.wall$over_represented_pvalue, method = "BH")

# Filter for significant terms (e.g., padj < 0.05)
enriched_GO <- GO.wall[GO.wall$padj < 0.05, ] # zero GO terms are significantly abundant 

# save goseq() results to a .csv file 
write_csv(GO.wall,
          file = "GOres_0.05_DAF_vs_CMT_July",
          path = "/work/clh162/OysterRNA24/DEG_analysis/GO/goseq_results/GOres_0.05_DAF_vs_CMT_July") 
```


## DAF vs. CMT August 

### Loading DESeq results + assigning old gene IDs 

```{r}
# results from August by site
res_0.05_DAF_vs_CMT_August <- read_csv("/work/clh162/OysterRNA24/DEG_analysis/DESeq2/res_0.05_DAF_vs_CMT_August")

# run a left_join() to get all the ENSEMBL gene names added to all the genes in this results set 
res_0.05_DAF_vs_CMT_August <- left_join(x = res_0.05_DAF_vs_CMT_August,
          y = RU17_RU25_gene_IDs,
          by = join_by(gene_id == RU25_gene)) # column names for RU25 genes are different between the two data sets 
res_0.05_DAF_vs_CMT_August

```

### Connect GO terms using ENSEMBL gene IDs

```{r}
# RESETING LENGTHS_VECTOR with the original gene_lengths file 
  # put length information into vector format  
lengths_vector <- gene_lengths$transcript_length
# add ensembl_gene_id names to the lengths data 
names(lengths_vector) <- gene_lengths$ensembl_gene_id

# Create DE gene vector where 1 = significant and 0 = not
DE_genes_vector <- as.integer(
  res_0.05_DAF_vs_CMT_August$padj < 0.05 &
  !is.na(res_0.05_DAF_vs_CMT_August$padj) # exclude DE genes that do not have a padj value 
)

# add ensembl_gene_id names to the lengths data 
names(DE_genes_vector) <- res_0.05_DAF_vs_CMT_August$gene



# Now restrict vectors to overlap with this DESeq result universe:
# create a list of common_genes shared between these two vectors 
common_genes <- intersect(names(DE_genes_vector), names(lengths_vector))

DE_genes_vector <- DE_genes_vector[common_genes]
lengths_vector <- lengths_vector[common_genes]
  # cut down from 21326 in DE_genes_vector and 39530 in lengths_vector to 16345 (the final # we expected from above)

```

### Investigation of differential expression + significance

```{r}

```

### Running goseq()

```{r}
# account for the fact that longer genes are easier to sequence/detect
pwf <- nullp(DEgenes = DE_genes_vector, bias.data = lengths_vector)

# run goseq
GO.wall <- goseq(pwf = pwf, 
                 gene2cat = gene2cat,
                 test.cats = ,
                 use_genes_without_cat = FALSE)
  # 3730 genes w/out GO terms -- excluded from p-value calculation
  # 5501 GO terms (rows)

# apply multiple testing correction method of Benjamini-Hochberg (BH) to control the False Discovery Rate (FDR) (aka limit type II errors)
GO.wall$padj <- p.adjust(GO.wall$over_represented_pvalue, method = "BH")

# Filter for significant terms (e.g., padj < 0.05)
enriched_GO <- GO.wall[GO.wall$padj < 0.05, ] # zero GO terms are significantly abundant 

# save goseq() results to a .csv file 
write_csv(GO.wall,
          file = "GOres_0.05_DAF_vs_CMT_August",
          path = "/work/clh162/OysterRNA24/DEG_analysis/GO/goseq_results/GOres_0.05_DAF_vs_CMT_August") 
```


## DAF - July vs. June 

### Loading DESeq results + assigning old gene IDs 

```{r}
# results between July and June (ref) at DAF
res_0.05_July_vs_June_DAF <- read_csv("/work/clh162/OysterRNA24/DEG_analysis/DESeq2/res_0.05_July_vs_June_DAF")

# run a left_join() to get all the ENSEMBL gene names added to all the genes in this results set 
res_0.05_July_vs_June_DAF <- left_join(x = res_0.05_July_vs_June_DAF,
          y = RU17_RU25_gene_IDs,
          by = join_by(gene_id == RU25_gene)) # column names for RU25 genes are different between the two data sets 
res_0.05_July_vs_June_DAF

```

### Connect GO terms using ENSEMBL gene IDs

```{r}
# RESETING LENGTHS_VECTOR with the original gene_lengths file 
  # put length information into vector format  
lengths_vector <- gene_lengths$transcript_length
# add ensembl_gene_id names to the lengths data 
names(lengths_vector) <- gene_lengths$ensembl_gene_id

# Create DE gene vector where 1 = significant and 0 = not
DE_genes_vector <- as.integer(
  res_0.05_July_vs_June_DAF$padj < 0.05 &
  !is.na(res_0.05_July_vs_June_DAF$padj) # exclude DE genes that do not have a padj value 
)

# add ensembl_gene_id names to the lengths data 
names(DE_genes_vector) <- res_0.05_July_vs_June_DAF$gene



# Now restrict vectors to overlap with this DESeq result universe:
# create a list of common_genes shared between these two vectors 
common_genes <- intersect(names(DE_genes_vector), names(lengths_vector))

DE_genes_vector <- DE_genes_vector[common_genes]
lengths_vector <- lengths_vector[common_genes]

```

### Investigation of differential expression + significance

```{r}

```

### Running goseq()

```{r}
# account for the fact that longer genes are easier to sequence/detect
pwf <- nullp(DEgenes = DE_genes_vector, bias.data = lengths_vector)

# run goseq
GO.wall <- goseq(pwf = pwf, 
                 gene2cat = gene2cat,
                 test.cats = ,
                 use_genes_without_cat = FALSE)
  # 3730 genes w/out GO terms -- excluded from p-value calculation
  # 5501 GO terms (rows)

# apply multiple testing correction method of Benjamini-Hochberg (BH) to control the False Discovery Rate (FDR) (aka limit type II errors)
GO.wall$padj <- p.adjust(GO.wall$over_represented_pvalue, method = "BH")

# Filter for significant terms (e.g., padj < 0.05)
enriched_GO <- GO.wall[GO.wall$padj < 0.05, ] # zero GO terms are significantly abundant 

# save goseq() results to a .csv file 
write_csv(GO.wall,
          file = "GOres_0.05_July_vs_June_DAF",
          path = "/work/clh162/OysterRNA24/DEG_analysis/GO/goseq_results/GOres_0.05_July_vs_June_DAF") 
```


## DAF - August vs. June 

### Loading DESeq results + assigning old gene IDs 

```{r}
# results between August and June (ref) at DAF
res_0.05_August_vs_June_DAF <- read_csv("/work/clh162/OysterRNA24/DEG_analysis/DESeq2/res_0.05_August_vs_June_DAF")

# run a left_join() to get all the ENSEMBL gene names added to all the genes in this results set 
res_0.05_August_vs_June_DAF <- left_join(x = res_0.05_August_vs_June_DAF,
          y = RU17_RU25_gene_IDs,
          by = join_by(gene_id == RU25_gene)) # column names for RU25 genes are different between the two data sets 
res_0.05_August_vs_June_DAF

```

### Connect GO terms using ENSEMBL gene IDs

```{r}
# RESETING LENGTHS_VECTOR with the original gene_lengths file 
  # put length information into vector format  
lengths_vector <- gene_lengths$transcript_length
# add ensembl_gene_id names to the lengths data 
names(lengths_vector) <- gene_lengths$ensembl_gene_id

# Create DE gene vector where 1 = significant and 0 = not
DE_genes_vector <- as.integer(
  res_0.05_August_vs_June_DAF$padj < 0.05 &
  !is.na(res_0.05_August_vs_June_DAF$padj) # exclude DE genes that do not have a padj value 
)

# add ensembl_gene_id names to the lengths data 
names(DE_genes_vector) <- res_0.05_August_vs_June_DAF$gene



# Now restrict vectors to overlap with this DESeq result universe:
# create a list of common_genes shared between these two vectors 
common_genes <- intersect(names(DE_genes_vector), names(lengths_vector))

DE_genes_vector <- DE_genes_vector[common_genes]
lengths_vector <- lengths_vector[common_genes]

```

### Investigation of differential expression + significance

```{r}

```

### Running goseq()

```{r}
# account for the fact that longer genes are easier to sequence/detect
pwf <- nullp(DEgenes = DE_genes_vector, bias.data = lengths_vector)

# run goseq
GO.wall <- goseq(pwf = pwf, 
                 gene2cat = gene2cat,
                 test.cats = ,
                 use_genes_without_cat = FALSE)
  # 3730 genes w/out GO terms -- excluded from p-value calculation
  # 5501 GO terms (rows)

# apply multiple testing correction method of Benjamini-Hochberg (BH) to control the False Discovery Rate (FDR) (aka limit type II errors)
GO.wall$padj <- p.adjust(GO.wall$over_represented_pvalue, method = "BH")

# Filter for significant terms (e.g., padj < 0.05)
enriched_GO <- GO.wall[GO.wall$padj < 0.05, ] # zero GO terms are significantly abundant 

# save goseq() results to a .csv file 
write_csv(GO.wall,
          file = "GOres_0.05_August_vs_June_DAF",
          path = "/work/clh162/OysterRNA24/DEG_analysis/GO/goseq_results/GOres_0.05_August_vs_June_DAF") 
```


## DAF - August vs. July 

### Loading DESeq results + assigning old gene IDs 

```{r}
# results between August and July at DAF
res_0.05_August_vs_July_DAF <- read_csv("/work/clh162/OysterRNA24/DEG_analysis/DESeq2/res_0.05_August_vs_July_DAF")

# run a left_join() to get all the ENSEMBL gene names added to all the genes in this results set 
res_0.05_August_vs_July_DAF <- left_join(x = res_0.05_August_vs_July_DAF,
          y = RU17_RU25_gene_IDs,
          by = join_by(gene_id == RU25_gene)) # column names for RU25 genes are different between the two data sets 
res_0.05_August_vs_July_DAF

```

### Connect GO terms using ENSEMBL gene IDs

```{r}
# RESETING LENGTHS_VECTOR with the original gene_lengths file 
  # put length information into vector format  
lengths_vector <- gene_lengths$transcript_length
# add ensembl_gene_id names to the lengths data 
names(lengths_vector) <- gene_lengths$ensembl_gene_id

# Create DE gene vector where 1 = significant and 0 = not
DE_genes_vector <- as.integer(
  res_0.05_August_vs_July_DAF$padj < 0.05 &
  !is.na(res_0.05_August_vs_July_DAF$padj) # exclude DE genes that do not have a padj value 
)

# add ensembl_gene_id names to the lengths data 
names(DE_genes_vector) <- res_0.05_August_vs_July_DAF$gene



# Now restrict vectors to overlap with this DESeq result universe:
# create a list of common_genes shared between these two vectors 
common_genes <- intersect(names(DE_genes_vector), names(lengths_vector))

DE_genes_vector <- DE_genes_vector[common_genes]
lengths_vector <- lengths_vector[common_genes]

```

### Investigation of differential expression + significance

```{r}

```

### Running goseq()

```{r}
# account for the fact that longer genes are easier to sequence/detect
pwf <- nullp(DEgenes = DE_genes_vector, bias.data = lengths_vector)

# run goseq
GO.wall <- goseq(pwf = pwf, 
                 gene2cat = gene2cat,
                 test.cats = ,
                 use_genes_without_cat = FALSE)
  # 3730 genes w/out GO terms -- excluded from p-value calculation
  # 5501 GO terms (rows)

# apply multiple testing correction method of Benjamini-Hochberg (BH) to control the False Discovery Rate (FDR) (aka limit type II errors)
GO.wall$padj <- p.adjust(GO.wall$over_represented_pvalue, method = "BH")

# Filter for significant terms (e.g., padj < 0.05)
enriched_GO <- GO.wall[GO.wall$padj < 0.05, ] # zero GO terms are significantly abundant 

# save goseq() results to a .csv file 
write_csv(GO.wall,
          file = "GOres_0.05_August_vs_July_DAF",
          path = "/work/clh162/OysterRNA24/DEG_analysis/GO/goseq_results/GOres_0.05_August_vs_July_DAF") 
```


## CMT - July vs. June

### Loading DESeq results + assigning old gene IDs 

```{r}
# results between July and June (ref) at CMT
res_0.05_July_vs_June_CMT <- read_csv("/work/clh162/OysterRNA24/DEG_analysis/DESeq2/res_0.05_July_vs_June_CMT")

# run a left_join() to get all the ENSEMBL gene names added to all the genes in this results set 
res_0.05_July_vs_June_CMT <- left_join(x = res_0.05_July_vs_June_CMT,
          y = RU17_RU25_gene_IDs,
          by = join_by(gene_id == RU25_gene)) # column names for RU25 genes are different between the two data sets 
res_0.05_July_vs_June_CMT

```

### Connect GO terms using ENSEMBL gene IDs

```{r}
# RESETING LENGTHS_VECTOR with the original gene_lengths file 
  # put length information into vector format  
lengths_vector <- gene_lengths$transcript_length
# add ensembl_gene_id names to the lengths data 
names(lengths_vector) <- gene_lengths$ensembl_gene_id

# Create DE gene vector where 1 = significant and 0 = not
DE_genes_vector <- as.integer(
  res_0.05_July_vs_June_CMT$padj < 0.05 &
  !is.na(res_0.05_July_vs_June_CMT$padj) # exclude DE genes that do not have a padj value 
)

# add ensembl_gene_id names to the lengths data 
names(DE_genes_vector) <- res_0.05_July_vs_June_CMT$gene



# Now restrict vectors to overlap with this DESeq result universe:
# create a list of common_genes shared between these two vectors 
common_genes <- intersect(names(DE_genes_vector), names(lengths_vector))

DE_genes_vector <- DE_genes_vector[common_genes]
lengths_vector <- lengths_vector[common_genes]

```

### Investigation of differential expression + significance

```{r}

```

### Running goseq()

```{r}
# account for the fact that longer genes are easier to sequence/detect
pwf <- nullp(DEgenes = DE_genes_vector, bias.data = lengths_vector)

# run goseq
GO.wall <- goseq(pwf = pwf, 
                 gene2cat = gene2cat,
                 test.cats = ,
                 use_genes_without_cat = FALSE)
  # 3730 genes w/out GO terms -- excluded from p-value calculation
  # 5501 GO terms (rows)

# apply multiple testing correction method of Benjamini-Hochberg (BH) to control the False Discovery Rate (FDR) (aka limit type II errors)
GO.wall$padj <- p.adjust(GO.wall$over_represented_pvalue, method = "BH")

# Filter for significant terms (e.g., padj < 0.05)
enriched_GO <- GO.wall[GO.wall$padj < 0.05, ] # zero GO terms are significantly abundant 

# save goseq() results to a .csv file 
write_csv(GO.wall,
          file = "GOres_0.05_July_vs_June_CMT",
          path = "/work/clh162/OysterRNA24/DEG_analysis/GO/goseq_results/GOres_0.05_July_vs_June_CMT")  
```


## CMT - August vs. June

### Loading DESeq results + assigning old gene IDs 

```{r}
# results between August and June (ref) at CMT
res_0.05_August_vs_June_CMT <- read_csv("/work/clh162/OysterRNA24/DEG_analysis/DESeq2/res_0.05_August_vs_June_CMT")

# run a left_join() to get all the ENSEMBL gene names added to all the genes in this results set 
res_0.05_August_vs_June_CMT <- left_join(x = res_0.05_August_vs_June_CMT,
          y = RU17_RU25_gene_IDs,
          by = join_by(gene_id == RU25_gene)) # column names for RU25 genes are different between the two data sets 
res_0.05_August_vs_June_CMT

```

### Connect GO terms using ENSEMBL gene IDs

```{r}
# RESETING LENGTHS_VECTOR with the original gene_lengths file 
  # put length information into vector format  
lengths_vector <- gene_lengths$transcript_length
# add ensembl_gene_id names to the lengths data 
names(lengths_vector) <- gene_lengths$ensembl_gene_id

# Create DE gene vector where 1 = significant and 0 = not
DE_genes_vector <- as.integer(
  res_0.05_August_vs_June_CMT$padj < 0.05 &
  !is.na(res_0.05_August_vs_June_CMT$padj) # exclude DE genes that do not have a padj value 
)

# add ensembl_gene_id names to the lengths data 
names(DE_genes_vector) <- res_0.05_August_vs_June_CMT$gene



# Now restrict vectors to overlap with this DESeq result universe:
# create a list of common_genes shared between these two vectors 
common_genes <- intersect(names(DE_genes_vector), names(lengths_vector))

DE_genes_vector <- DE_genes_vector[common_genes]
lengths_vector <- lengths_vector[common_genes]

```

### Investigation of differential expression + significance

```{r}

```

### Running goseq()

```{r}
# account for the fact that longer genes are easier to sequence/detect
pwf <- nullp(DEgenes = DE_genes_vector, bias.data = lengths_vector)

# run goseq
GO.wall <- goseq(pwf = pwf, 
                 gene2cat = gene2cat,
                 test.cats = ,
                 use_genes_without_cat = FALSE)
  # 3730 genes w/out GO terms -- excluded from p-value calculation
  # 5501 GO terms (rows)

# apply multiple testing correction method of Benjamini-Hochberg (BH) to control the False Discovery Rate (FDR) (aka limit type II errors)
GO.wall$padj <- p.adjust(GO.wall$over_represented_pvalue, method = "BH")

# Filter for significant terms (e.g., padj < 0.05)
enriched_GO <- GO.wall[GO.wall$padj < 0.05, ] # zero GO terms are significantly abundant 

# save goseq() results to a .csv file 
write_csv(GO.wall,
          file = "GOres_0.05_August_vs_June_CMT",
          path = "/work/clh162/OysterRNA24/DEG_analysis/GO/goseq_results/GOres_0.05_August_vs_June_CMT") 
```


## CMT - August vs. July

### Loading DESeq results + assigning old gene IDs 

```{r}
# results between August and July at CMT
res_0.05_August_vs_July_CMT <- read_csv("/work/clh162/OysterRNA24/DEG_analysis/DESeq2/res_0.05_August_vs_July_CMT")

# run a left_join() to get all the ENSEMBL gene names added to all the genes in this results set 
res_0.05_August_vs_July_CMT <- left_join(x = res_0.05_August_vs_July_CMT,
          y = RU17_RU25_gene_IDs,
          by = join_by(gene_id == RU25_gene)) # column names for RU25 genes are different between the two data sets 
res_0.05_August_vs_July_CMT

```

### Connect GO terms using ENSEMBL gene IDs

```{r}
# RESETING LENGTHS_VECTOR with the original gene_lengths file 
  # put length information into vector format  
lengths_vector <- gene_lengths$transcript_length
# add ensembl_gene_id names to the lengths data 
names(lengths_vector) <- gene_lengths$ensembl_gene_id

# Create DE gene vector where 1 = significant and 0 = not
DE_genes_vector <- as.integer(
  res_0.05_August_vs_July_CMT$padj < 0.05 &
  !is.na(res_0.05_August_vs_July_CMT$padj) # exclude DE genes that do not have a padj value 
)

# add ensembl_gene_id names to the lengths data 
names(DE_genes_vector) <- res_0.05_August_vs_July_CMT$gene



# Now restrict vectors to overlap with this DESeq result universe:
# create a list of common_genes shared between these two vectors 
common_genes <- intersect(names(DE_genes_vector), names(lengths_vector))

DE_genes_vector <- DE_genes_vector[common_genes]
lengths_vector <- lengths_vector[common_genes]

```

### Investigation of differential expression + significance

```{r}

```

### Running goseq()

```{r}
# account for the fact that longer genes are easier to sequence/detect
pwf <- nullp(DEgenes = DE_genes_vector, bias.data = lengths_vector)

# run goseq
GO.wall <- goseq(pwf = pwf, 
                 gene2cat = gene2cat,
                 test.cats = ,
                 use_genes_without_cat = FALSE)
  # 3730 genes w/out GO terms -- excluded from p-value calculation
  # 5501 GO terms (rows)

# apply multiple testing correction method of Benjamini-Hochberg (BH) to control the False Discovery Rate (FDR) (aka limit type II errors)
GO.wall$padj <- p.adjust(GO.wall$over_represented_pvalue, method = "BH")

# Filter for significant terms (e.g., padj < 0.05)
enriched_GO <- GO.wall[GO.wall$padj < 0.05, ] # zero GO terms are significantly abundant 

# save goseq() results to a .csv file 
write_csv(GO.wall,
          file = "GOres_0.05_August_vs_July_CMT",
          path = "/work/clh162/OysterRNA24/DEG_analysis/GO/goseq_results/GOres_0.05_August_vs_July_CMT") 

# save significant GO terms to a .csv file 
write_csv(enriched_GO,
          file = "enriched_GO_0.05_August_vs_July_CMT",
          path = "/work/clh162/OysterRNA24/DEG_analysis/GO/goseq_results/enriched_GO_0.05_August_vs_July_CMT")
```

# Reading in results of goseq to dataframes 

```{r}
### Site comparisons at each timepoint ###
# DAF vs. CMT June 
GOres_0.05_DAF_vs_CMT_June <- read_csv("/work/clh162/OysterRNA24/DEG_analysis/GO/goseq_results/GOres_0.05_DAF_vs_CMT_June")
# DAF vs. CMT July 
GOres_0.05_DAF_vs_CMT_July <- read_csv("/work/clh162/OysterRNA24/DEG_analysis/GO/goseq_results/GOres_0.05_DAF_vs_CMT_July")
# DAF vs. CMT Aug 
GOres_0.05_DAF_vs_CMT_August <- read_csv("/work/clh162/OysterRNA24/DEG_analysis/GO/goseq_results/GOres_0.05_DAF_vs_CMT_August")

### DAF timepoint comparisons ### 
# DAF - July vs. June 
GOres_0.05_July_vs_June_DAF <- read_csv("/work/clh162/OysterRNA24/DEG_analysis/GO/goseq_results/GOres_0.05_July_vs_June_DAF")
# DAF - Aug vs. June
GOres_0.05_August_vs_June_DAF <- read_csv("/work/clh162/OysterRNA24/DEG_analysis/GO/goseq_results/GOres_0.05_August_vs_June_DAF")
# DAF - Aug vs. July 
GOres_0.05_August_vs_July_DAF <- read_csv("/work/clh162/OysterRNA24/DEG_analysis/GO/goseq_results/GOres_0.05_August_vs_July_DAF")

### CMT timepoint comparisons ###
# CMT - July vs. June 
GOres_0.05_July_vs_June_CMT <- read_csv("/work/clh162/OysterRNA24/DEG_analysis/GO/goseq_results/GOres_0.05_July_vs_June_CMT")
# CMT - Aug vs. June
GOres_0.05_August_vs_June_CMT <- read_csv("/work/clh162/OysterRNA24/DEG_analysis/GO/goseq_results/GOres_0.05_August_vs_June_CMT")
# CMT - Aug vs. July 
GOres_0.05_August_vs_July_CMT <- read_csv("/work/clh162/OysterRNA24/DEG_analysis/GO/goseq_results/GOres_0.05_August_vs_July_CMT")

```











# NOTHING BELOW THIS LINE RUN -- other potential pipeline(?)

Let's focus on running our GO analysis only on a subset of significantly expressed genes. First to get the significantly differentially expressed genes, we can subset from our results dataframe from the dds. 

```{r}
sign_DEG <- na.omit(results)
sign_DEG <- sign_DEG[sign_DEG$padj < 0.05 & sign_DEG$baseMean > 50] # need to decide what threshold values we want to use for defining significance
```

## GO analysis for positively differentially expressed genes that have a value above 0.5 for log2FoldChange 

First, we can subset our significant DEG dataset (sign_DEG) to target this group. 

```{r}
sign_DEG[sign_DEG$log2FoldChange > 0.5, ]
```

Then we need to pull out a list of the gene names from this subset.

```{r}
rownames(sign_DEG[sign_DEG$log2FoldChange > 0.5, ])
# save this list 
list <- rownames(sign_DEG[sign_DEG$log2FoldChange > 0.5, ])
```

Using this list we can now run GO analysis

```{r}
GO_results <- clusterProfiler::enrichGO(genes = genes_to_test,
                                        OrgDb = ,
                                        keyType = )
```
