---
title: "WGCNA analysis"
author: "Callie Hundley"
format: 
  html:
    df-print: paged 
    code-fold: true 
    code-tools: true
editor: visual
editor_options: 
  chunk_output_type: console
---

# Source citation

The following informational materials were used in constructing the pipeline outlined in this quarto document:

-   [Bioinformatics workbook: Network analysis with WGCNA](https://bioinformaticsworkbook.org/tutorials/wgcna.html#gsc.tab=0)
-   [WGCNA Step-by-step Tutorial - Part 1](https://youtu.be/gYE59uEMXT4?si=ll_pBKd00fDDIsbz)
-   [WGCNA Step-by-step Tutorial - Part 2](https://youtu.be/mzXIxjPr_Mc?si=JeVZEtXsFYwCuCdT)

# Weighted Gene Co-expression Network Analysis (WGCNA)

A subsequent reasonable step in our analysis is to run a WGCNA. When we run this on our significantly differentially expressed genes, it will create clusters/modules of similarly expressed (or co-expressed) genes. This can be helpful as it will direct us into interesting genes to look at for additional analysis like gene ontology (GO) classification, instead of starting with our thousands of initial genes.

What we need to do here:

-   Construct a correlation network, calculate soft thresholds, and group genes into modules.
-   Correlate with Traits: Identify which gene modules correlate with your experimental conditions (in our case location, month, or environmental conditions) or phenotype (i.e. mortality, but we don't have much of that in these samples).

## Load packages

```{r, warning=FALSE}
#| echo: FALSE
# Install the Bioconductor manager if you don't have it
#if (!requireNamespace("BiocManager", quietly = TRUE))
    #install.packages("BiocManager")
library(BiocManager)
# Install required Bioconductor dependencies
#BiocManager::install(c("BiocGenerics", "AnnotationDbi", "GO.db", "preprocessCore", "impute"))

#install.packages("WGCNA")
library(WGCNA)
library(tidyverse)
library(DESeq2)
#devtools::install_github("kevinblighe/CorLevelPlot")
library(CorLevelPlot)
library(readr)
library(readxl)
library(gridExtra)
library(patchwork)
```

## Load and prepare data

### Gene expression data

```{r}
# Reload saved DESeq2 dataset object 
# if using unfiltered DESeqResults object
# dds <- readRDS("/work/clh162/OysterRNA24/DEG_analysis/DESeq2/count_dds.rds")

# if using filtered DESeqResults object  
dds_filtered <- readRDS("/work/clh162/OysterRNA24/DEG_analysis/DESeq2/DESeqResults_Objects/DESeqResults_dds_filtered.rds")
```

### Metadata (traits)

```{r}
coldata <- read_excel("/work/clh162/OysterRNA24/xRNA for BC Summer 2024.xlsx",
                      col_types = c("text", "text", "text", "text", "text", "date", "date", "numeric", 
                                    "numeric", "numeric", "numeric"),
                      skip = 2) %>% #skipping first two rows of .xls file
            column_to_rownames(var = "Sample_ID") %>% #assigning the sample names as the row names 
            mutate(month_collected = month(Date_Collected, 
                                           label = TRUE, 
                                           abbr = FALSE)) # creating streamlined column for graphing/analysis by date
```

### Environmental data (traits)

```{r}
# Load environmental_7d_avg_data
env_data <- read_csv("/work/clh162/OysterRNA24/DEG_analysis/WGCNA/envir_traits_7day_summary_2024.csv")

colnames(env_data)

env_mat <- env_data[, c("Farm", "Date_Collected", "temperature_7d_observed_mean", "salinity_7d_observed_mean", "conductivity_7d_observed_mean", "do_7d_observed_mean", "precipitation_7d_observed_mean", "pH_7d_observed_mean", "pH_7d_interpolated_mean")]

# need to do this? 
# rownames(env_data) <- rownames(col_data)
```

## Quality control check -- Outliers

We can use the function `goodSamplesGenes()` to detect outliers in our count data. Requires the input data have samples assigned to rows position and genes assigned to the columns position. Since our count data is in the reverse format in the dds object, we can transpose our data using the `t()` function which flips the rows and columns of the dataset.

```{r}
goodSamplesGenes(t())
```

## Normalize count data

We will use normalized gene counts for our WGCNA analysis. We can use variance stabilization to transform our counts to normalized counts. Again we also need to get these normalized counts into the format where samples are the rows and genes are in the columns position.

```{r}
dds_norm <- vst(dds_filtered, blind = TRUE) # default for blind = is TRUE -- leave as is 

# the normalized counts are stored in the assay() of the dds_norm object 
head(assay(dds_norm))

# save normalized data  as matrix 
norm.counts <- assay(dds_norm) 
```

## Pull top DEG for WGCNA 

Use the top 5% of filtered genes by variance, matching the scale of the earlier exploratory WGCNA.

```{r}
gene_variance <- matrixStats::rowVars(norm.counts)
names(gene_variance) <- rownames(norm.counts)

number_of_wgcna_genes <- ceiling(length(gene_variance) * 0.10)
number_of_wgcna_genes
selected_wgcna_genes <- names(
  sort(gene_variance, decreasing = TRUE)
)[seq_len(number_of_wgcna_genes)]

datExpr_top <- t(
  norm.counts[selected_wgcna_genes, , drop = FALSE]
)

```


## Construct weighted correlation matrix

### Determine soft threshold

First we need to pick the power level which will produce a weighted correlation matrix that is most similar to a scale free network topology.

```{r}
# create a vector of soft-thresholding powers for selection 
power <- c(c(1:10), seq(from = 12, to = 20, by = 2))

# use network topology analysis function for each power level
sft <- pickSoftThreshold(data = datExpr_top,
                  powerVector = power,
                  networkType = "signed",
                  verbose = 5)

# pull out data for deciding which power produces highest similarity to a scale free topology
sft.data <- data.frame(sft$fitIndices)  
  # will use two values from here: 
    # SFT.R.sq
    # mean.k = mean connectivity 
      # want to pick the power that provides the maximum R.sq value and minimum mean.k

# create visualization to help pick which power level to use 
# plot for SFT.R.sq
R.sq.plot <- ggplot(data = sft.data, 
                    aes(x = Power, y = SFT.R.sq, label = Power)) +
  geom_point() + 
  geom_text(nudge_y = 0.1) +
  geom_hline(yintercept = 0.8, color = "red") +
  labs(x = "Power",
       y = "Scale free topology model fit, signed R^2") + 
  theme_classic()

# plot for mean.k
mean.k.plot <- ggplot(data = sft.data, 
                      aes(x = Power, y = mean.k., label = Power)) + 
  geom_point() + 
  geom_text(nudge_y = 0.1) +
  labs(x = "Power",
       y = "Mean Connectivity") + 
  theme_classic()

# display plots together 
grid.arrange(R.sq.plot, mean.k.plot, nrow = 2)
```

Looking at these plots together, ideally we should select a power value that has a R\^2 value greater than 0.8 (above the red line we added) but not excessively large as well as minimal mean connectivity.

Next we need to convert our matrix of normalized counts to numeric format to run the `blockwiseModules()` function.

```{r}
datExpr_top[] <- sapply(datExpr_top, as.numeric)

# set the assigned power we picked above
soft_power <- 7

temp_cor <- cor

cor <- WGCNA::cor # Force it to use WGCNA cor function (fix a namespace conflict issue)

# memory estimate w.r.t. blocksize 
bwnetwk <- blockwiseModules(datExpr = datExpr_top, # input 
                  # -- Adjacency Function -- #
                 maxBlockSize = 5000, # how many genes maximum you want to be included in each module, default is 5,000
                 minModuleSize = 50,
                 TOMType = "signed",
                 power = soft_power, # power value we just assigned above
                 mergeCutHeight = 0.25, # will merge/cluster seperate modules that have > 75% correlations together  
                 numericLabels = FALSE, # want the module eigengene labels to be colors instead of the # of genes 
                 randomSeed = 1234, # set randomSeed for reproducability
                 verbose = 3) 

cor <- temp_cor

# Save blockwise network object to RDS file
saveRDS(bwnetwk, file = "/work/clh162/OysterRNA24/DEG_analysis/WGCNA/wgcna_results.rds")
```

## Identify gene modules

Let's look at the resulting **Module Eigengenes** that are returned from the `blockwiseModules()` function. Based on the dissimilarity values of these modules (distance between the module eigengenes), we can cluster/merge similar modules together.

```{r}
# return the module eigengene information from the variable storing our blockwiseModules() function 
module_eigengenes <- bwnetwk$MEs

# Extract module assignments from the completed WGCNA network
module_colors <- bwnetwk$colors

# Convert to a simple character vector if WGCNA returned a list-like object
if (is.list(module_colors)) {
  module_colors <- unlist(
    module_colors,
    recursive = TRUE,
    use.names = FALSE
  )
}

module_colors <- as.character(
  module_colors
)

# Each WGCNA gene should have exactly one module assignment
stopifnot(
  length(module_colors) == ncol(datExpr_top),
  !anyNA(module_colors)
)

names(module_colors) <- colnames(
  datExpr_top
)

# Create module-size summary
module_size_table <- tibble::tibble(
  module = module_colors
) %>%
  dplyr::group_by(module) %>%
  dplyr::summarise(
    number_of_genes = dplyr::n(),
    .groups = "drop"
  ) %>%
  dplyr::arrange(
    dplyr::desc(number_of_genes)
  )

# Print the top of the Module Eigengenes results 
head(module_eigengenes)

# return number of genes in each module
table(bwnetwk$colors)

# visualize clustering of modules with a dendrogram both before and after merging modules (based on 75% similarity)  
plotDendroAndColors(bwnetwk$dendrograms, 
                    cbind(bwnetwk$unmergedColors, bwnetwk$colors),
                    c("unmerged", "merged"),
                    dendroLabels = FALSE,
                    addGuide = TRUE,
                    hang = 0.03,
                    guideHang = 0.05
                    )
#### COME BACK TO THIS + FIX dendrogram ########
```

Because we have fewer modules (colors) post merging, we will use the merged modules to perform further downstream analysis.

## Additional analyses

### Calculate Module Eigengenes

```{r}
module_eigengenes <- moduleEigengenes(
  datExpr_top,
  colors = module_colors
)$eigengenes

module_eigengenes <- orderMEs(module_eigengenes)

stopifnot(
  identical(
    rownames(module_eigengenes),
    rownames(coldata)
  )
) # currently FALSE because coldata rownames have "_Gill" on the end of them 

dim(module_eigengenes)
names(module_eigengenes)

# integrate sample metadata with module eigengenes 
sample_level_eigengenes <- bind_cols(
  coldata %>%
    rownames_to_column(var = "metadata_rowname") %>%
    select(-metadata_rowname),
  as_tibble(module_eigengenes)
)

# save sample level eigengene data
readr::write_csv(
  sample_level_eigengenes,
  file = "top0.10DEG_WGCNA_sample_level_eigengenes.csv",
  path = "/work/clh162/OysterRNA24/DEG_analysis/WGCNA/top0.10DEG_WGCNA_sample_level_eigengenes.csv"
)
```

### Collapse eigengenes to the six tissue collection events 

This is important so we can later pair this data with our environmental means 

```{r}
# isolate eigengene column names 
eigengene_columns <- grep(
  "^ME",
  names(sample_level_eigengenes),
  value = TRUE
)

event_level_eigengenes <- sample_level_eigengenes %>%
  group_by(Location, month_collected, Date_Collected) %>%
  summarise(
    across(all_of(eigengene_columns), mean),
    RNAseq_samples = n(),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    event_id = paste0(
      Location,
      "_",
      format(Date_Collected, "%Y-%m-%d")
    )
  ) %>% 
  dplyr::relocate(
    event_id, 
    .before = Location
  ) %>% 
  arrange(Date_Collected, Location)

# check to make sure biological replicates are being averaged correctly  
stopifnot(
  nrow(event_level_eigengenes) == 6,
  all(event_level_eigengenes$RNAseq_samples == 3)
)

# save collection event level eigengene data
readr::write_csv(
  event_level_eigengenes,
  file = "top0.10DEG_WGCNA_event_level_eigengenes.csv",
  path = "/work/clh162/OysterRNA24/DEG_analysis/WGCNA/top0.10DEG_WGCNA_event_level_eigengenes.csv"
)

```


### Correlate modules with pheontypes/traits

Now that we have created modules and merged similar modules, the next good step of analysis would be to see how the modules relate to traits in our samples. In other words, **which modules have significant association with a trait of interest?**

In order to investigate this question, we have to binarize any categorical trait variables that we want to check their associations with to our module eigengenes. Remember that our trait (metadata) associated with our samples is stored in the object `coldata`.

#### Combine eigengene values + traits into data frame 

```{r}
# Create the finalized event-level environmental table (from env_data)
environmental_events_final <-
  env_data %>%
  dplyr::mutate(
    Location = Farm,
    Date_Collected = as.Date(Date_Collected),
    event_id = paste0(
      Farm,
      "_",
      format(Date_Collected, "%Y-%m-%d")
    )
  ) %>%
  select(!c(Farm)) %>% 
  dplyr::relocate(
    event_id, Location,
    .before = Date_Collected
  )

# Preview the result
environmental_events_final

# Combine environmental event data with event level eigengene data 
wgcna_traits <- event_level_eigengenes %>%
  left_join(
    environmental_events_final, # COME BACK AND MAKE THIS CLEANER BY REMOVING SOME COLUMNS THAT ARE NOT NEEDED!
    by = c("Location", "month_collected", "Date_Collected")
  )

# check to make sure retained all data 
stopifnot(
  nrow(wgcna_traits) == 6,
  n_distinct(wgcna_traits$event_id) == 6
)

```

Now that we have our module eigengene data and sample metadata + environmental means all in the same table (`wgcna_traits`) we can build an expanded trait matrix that has all of our traits in a quantitative or binarized (for categorical traits) format.  

#### Format quantitative/binarized trait matrix

We will only retain our traits of interest that we want to test for correlations with our WGCNA module eigengene information. Here, we also save our `event_id` as the rownames of this matrix so that this trait matrix can cross-talk with our eigengene matrix.  

```{r}

# Create the trait data table
wgcna_traits_binarized <- wgcna_traits %>%
  dplyr::mutate(
    # collapse the pH columns into one column (ph_7d_filled) - to make correlation/calculations easier 
      # just remember that the CMT Aug collected sample has interpolated 7d mean + all others are observed 
    pH_7d_filled_mean = coalesce(pH_7d_observed_mean, pH_7d_interpolated_mean)
  ) %>% 
  dplyr::transmute( #transmute automatically only keeps the rows that you mention in the command (drops all others)
    event_id = as.character(event_id),

    # Month
    month_Jun = as.integer(month_collected == "June"),
    month_Jul = as.integer(month_collected == "July"),
    month_Aug = as.integer(month_collected == "August"),
    
    # Location 
    location_DAF = as.integer(Location == "DAF"),
    
    # Environmental traits 
    temperature_7d_observed_mean,
    do_7d_observed_mean,
    salinity_7d_observed_mean,
    conductivity_7d_observed_mean,
    pH_7d_filled_mean,
    precipitation_7d_observed_mean,

  ) 

# Convert only numeric trait columns into the matrix form of this table
wgcna_traits_binarized_matrix <- wgcna_traits_binarized %>%
  dplyr::select(-event_id) %>%
  as.matrix()

# Assign event IDs AFTER conversion to matrix
rownames(wgcna_traits_binarized_matrix) <-
  wgcna_traits_binarized$event_id

# We now have our eigengene_matrix and trait_matrix with rownames that match between the two (for communication purposes)
# eigengene_matrix

# trait_matrix
head(wgcna_traits_binarized_matrix)
```

#### Format eigengene value matrix 

Here, we also save our `event_id` as the rownames of this matrix so that this eigengene matrix can cross-talk with our trait matrix.  

```{r}
wgcna_eigengene_matrix <-
  event_level_eigengenes %>%
  dplyr::select(
    dplyr::all_of(eigengene_columns) # list defined earlier: the names of the columns associated with eigengene data in event_level_eigengenes 
  ) %>%
  as.matrix()

rownames(wgcna_eigengene_matrix) <-
  as.character(
    event_level_eigengenes$event_id
  )
```

Let's make sure that our two matrices have matching values and are ready to plug into our function:

```{r}
# check the row names of each 
rownames(wgcna_eigengene_matrix)
rownames(wgcna_traits_binarized_matrix)

# check that row names (event_id) are identical 
identical(
  rownames(wgcna_eigengene_matrix),
  rownames(wgcna_traits_binarized_matrix)
) # TRUE

# check dimensions 
dim(wgcna_eigengene_matrix) # 6 x 5
dim(wgcna_traits_binarized_matrix) # 6 x 10

# make sure no values are missing 
sum(is.na(wgcna_eigengene_matrix))
colSums(is.na(wgcna_traits_binarized_matrix))
```


#### Create function to calculate correlations/p-values

Next we can create a **function** to calculate correlation + p-values between module eigengenes and traits 

```{r}
pairwise_correlation_results <- function(eigengene_matrix, trait_matrix) {
  # requires two inputs:
    #1. an eigengene matrix (that contains the eigengene values for your WGCNA modules + some sample identifier)
    #2. a trait_matrix (that contains quantitative and/or binarized traits associated with ^some sample identifier)
  output <- vector(
    "list",
    ncol(eigengene_matrix) * ncol(trait_matrix)
  )

  result_number <- 1

  for (module_name in colnames(eigengene_matrix)) {
    for (trait_name in colnames(trait_matrix)) {
      module_values <- eigengene_matrix[, module_name]
      trait_values <- trait_matrix[, trait_name]

      complete_rows <- complete.cases(module_values, trait_values)
      n_complete <- sum(complete_rows)

      if (
        n_complete >= 3 &&
        sd(module_values[complete_rows]) > 0 &&
        sd(trait_values[complete_rows]) > 0
      ) {
        test_result <- cor.test(
          module_values[complete_rows],
          trait_values[complete_rows],
          method = "pearson"
        )

        correlation_value <- unname(test_result$estimate)
        p_value <- test_result$p.value
      } else {
        correlation_value <- NA_real_
        p_value <- NA_real_
      }

      output[[result_number]] <- tibble(
        module = module_name,
        trait = trait_name,
        correlation = correlation_value,
        n_events = n_complete,
        p_value = p_value
      )

      result_number <- result_number + 1
    }
  }

  bind_rows(output) %>%
    mutate(
      FDR = p.adjust(p_value, method = "BH"),
      absolute_correlation = abs(correlation)
    )
}
```

#### Run function! 

Finally, using this function let's calculate these correlations + p-values 

```{r}

module_trait_results <- pairwise_correlation_results(
  wgcna_eigengene_matrix,
  wgcna_traits_binarized_matrix
) %>%
  arrange(FDR, p_value, desc(absolute_correlation))

# save as csv 
readr::write_csv(
  module_trait_results,
  file = "top0.10DEG_WGCNA_module_trait_correlations.csv",
  path = "/work/clh162/OysterRNA24/DEG_analysis/WGCNA/top0.10DEG_WGCNA_module_trait_correlations.csv"
)

module_trait_results %>%
  slice_head(n = 25)
```

### Create WGCNA heatmap w/module + trait correlations 

```{r}
# Eigengene matrix: wgcna_eigengene_matrix
# Trait matrix: wgcna_traits_binarized_matrix
# Results table (with correlations/FDR,p-values): module_trait_results

# ------------------------------------------------------------
# Preserve the module and environmental-trait ordering
# ------------------------------------------------------------
module_order <- colnames(wgcna_eigengene_matrix)
trait_order <- colnames(wgcna_traits_binarized_matrix)

# create a vector of how to graph/rename traits in heatmap display
trait_display_names <- c(
  temperature_7d_observed_mean = "Temperature",
  do_7d_observed_mean = "Dissolved oxygen",
  salinity_7d_observed_mean = "Salinity",
  conductivity_7d_observed_mean = "Conductivity",
  pH_7d_filled_mean = "pH\n(linear interp for CMT_Aug)",
  precipitation_7d_observed_mean = "Precipitation",
  month_Jun = "June",
  month_Jul = "July",
  month_Aug = "August",
  location_DAF = "Location\nDAF vs CMT"
)

# ------------------------------------------------------------
# Prepare module-trait results
# ------------------------------------------------------------

heatmap_plot_data <- module_trait_results %>%
  mutate(
    # Reverse levels so the first module appears at the top
    module = factor(
      module,
      levels = rev(module_order)
    ),

    trait = factor(
      trait,
      levels = trait_order
    ),

    # Label only nominally significant relationships (p < 0.05)
    significant = !is.na(p_value) & p_value < 0.05,

    # distinguish display of different p-values
    p_display = case_when(
      is.na(p_value) ~ "NA",
      p_value < 0.001 ~ "<0.001",
      TRUE ~ formatC(
        p_value,
        format = "f",
        digits = 3
      )
    ),

    # information to print within each heatmap cell 
    cell_label = if_else(
      significant,
      paste0(
        formatC(
          correlation,
          format = "f",
          digits = 2
        ),
        "\n(p=",
        p_display,
        ")"
      ),
      ""
    )
  )

# ------------------------------------------------------------
# create correlation heatmap 
# ------------------------------------------------------------

correlation_heatmap <- ggplot(
  heatmap_plot_data,
  aes(
    x = trait,
    y = module,
    fill = correlation
  )
) +
  geom_tile(
    color = "grey85",
    linewidth = 0.5
  ) +
  geom_text(
    aes(label = cell_label), # see above set-up
    size = 3,
    fontface = "bold",
    lineheight = 0.9
  ) +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#D73027",
    midpoint = 0,
    limits = c(-1, 1),
    name = "Correlation"
  ) +
  scale_x_discrete(
    labels = trait_display_names,
    drop = FALSE
  ) +
  scale_y_discrete(
    drop = FALSE
  ) +
  labs(
    title = "WGCNA module-trait correlations",
    subtitle = "Labels shown significance only for nominal p < 0.05",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),

    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      vjust = 1,
      size = 10
    ),

    axis.text.y = element_text(
      size = 11
    ),

    plot.title = element_text(
      face = "bold"
    ),

    legend.position = "left",

    plot.margin = margin(
      t = 5,
      r = 5,
      b = 5,
      l = 5
    )
  )
correlation_heatmap

# ------------------------------------------------------------
# create gene count plot (# of genes in each module)
# ------------------------------------------------------------

# organize gene count in each module data to plot 
module_count_plot_data <- module_size_table %>% #created above, the module color + # of genes in each 
  # filter(module != "grey") %>% # getting rid of a certain module
  mutate(
    module_ME = paste0("ME", module),

    module_ME = factor(
      module_ME,
      levels = rev(module_order)
    )
  ) %>%
  filter(!is.na(module_ME)) # get rid of na values 

# create bar chart 
gene_count_plot <- ggplot(
  module_count_plot_data,
  aes(
    x = number_of_genes,
    y = module_ME
  )
) +
  geom_col(
    aes(fill = module),
    width = 0.72,
    show.legend = FALSE
  ) +
  scale_fill_identity() +
  geom_text(
    aes(
      label = scales::comma(number_of_genes)
    ),
    hjust = -0.15,
    size = 4
  ) +
  scale_y_discrete(
    drop = FALSE
  ) +
  scale_x_continuous(
    limits = c(
      0,
      max(module_count_plot_data$number_of_genes) * 1.4
    ),
    expand = expansion(mult = c(0, 0))
  ) +
  labs(
    subtitle = "Number of genes",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),

    plot.title = element_text(
      face = "bold",
      hjust = 0.5,
      size = 12
    ),

    axis.text = element_blank(),
    axis.ticks = element_blank(),

    plot.margin = margin(
      t = 5,
      r = 10,
      b = 5,
      l = 0
    )
  )


# ------------------------------------------------------------
# combine correlation heatmap + gene count plot into one figure 
# ------------------------------------------------------------

final_wgcna_plot <-
  correlation_heatmap +
  gene_count_plot +
  plot_layout(
    widths = c(5.5, 1.4)
  )

final_wgcna_plot


# save WGCNA heatmap + gene count plot 

ggsave(
  plot = final_wgcna_plot,
  filename = "heatmap_top0.10DEG_sig_cor_and_gene_counts.png",
  path = "/work/clh162/OysterRNA24/DEG_analysis/WGCNA",
  width = 10,
  height = 5,
  units = "in",
  dpi = 800,
  bg = "white"
)
```


#-------------EXTRA NOTES FROM BEFORE------------------------------------

#### Calculate correlation + p-values

Now we can combine our binarized trait data with the gene module results and calculate the correlation and p-values.

```{r}
# Define numbers of samples and genes 
nSamples <- nrow(norm.counts)
nGenes <- ncol(norm.counts)

# calculate correlation between module eigengenes and traits using the cor() function
module.trait.corr <- cor(module_eigengenes, traits, use = "p")
module.trait.corr.pvals <- corPvalueStudent(module.trait.corr, nSamples)

```

#### Create heatmap

Finally we can visualize these module-trait associations using a heatmap.

```{r}
# create dataframe that combines module_eigengenes and trait data into one place 
heatmap.data <- merge(module_eigengenes, traits, by = "row.names")

heatmap.data <- heatmap.data %>% 
  column_to_rownames(var = "row.names") # make sure the first column is read as the row names, not as a column (CorLevelPlot() will be mad because it only takes categorical variables that are binarized!)

# create heatmap
CorLevelPlot(data = heatmap.data,
             x = names(heatmap.data)[#:#], # columns on x-axis: trait values 
             y = names(heatmap.data)[#:#], # columns on y-axis: modules 
             col = c("blue1", "skyblue", "white", "pink", "red"))
```

After identifying what modules are highly associated with different traits, we can investigate the genes that are present within these modules to try to understand mechanistically what might be driving these associations.

First we can extract the genes associated with each module.

```{r}
module.gene.mapping <- as.data.frame(bwnetwk$colors)

# filter this assignment list of genes for a given module of interest 
module.gene.mapping %>% 
  filter(bwnetwk$colors == "turquoise") %>% # specify module 
  rownames()
```

### Identify driver genes

#### Calculating module membership

We can identify highly connected "hub" genes within a single module of interest by calculating the **module membership** and associated p-values. The module membership/intramodular connectivity is calculated as the correlation of the model eigengenes and the (normalized) gene expression profile. In other words, module membership quantifies the similarity of all genes expressed to every module.

```{r}
# calculate the module membership measures and the associated p-values 
module.membership.measure <- cor(x = module_eigengenes, 
                                 y = norm.counts, 
                                 use = "p") # using Person correlation
module.membership.measure.pvals <- corPvalueStudent(module.membership.measure, nSamples)
```

By looking at the module membership measures and their p-values, we can determine which genes have significantly high membership (importance) within a given module of interest. We can subsequently extract those genes to further look into them. \#### Calculating gene significance

We can also identify genes highly associated with a given trait of interest by calculating **gene significance** and associated p-values. Gene significance is calculated as the correlation of the (normalized) gene expression profile and trait of interest.

```{r}
gene.signf.measure <- cor(x = norm.counts,
                                 y = traits$#PICKONE_HERE,
                                 use = "p") # using Person correlation
gene.signf.measure.pvals <- corPvalueStudent(gene.signf.measure.pvals, nSamples)

# take a look at resulting correlations 
gene.signf.measure.pvals %>% 
  as.data.frame() %>% 
  arrange(p-valuecolumn) %>% # replace with the name of the column containing p-values in the object -- this line displays p-values from smallest to largest
    head(25) # return top 25 genes
```

We can subsequently extract these gene names and perform downstream analysis like looking at their gene symbols and run GO enrichment.

# Note on methods to associate module eigengenes (continuous variable) with sample traits (categorical variables)

There are multiple methods you can use to test association of these two datasets.

But regardless of which method you use (e.g. running a student t-test, pearson correlation, or using a linear model) ultimately you are going to get the same list of modules to be significantly associated with the trait of interest.
