---
title: "wgcna"
output: html_document
date: "2026-06-12"
---

# Setup
```{r}
# load in libraries
library(BiocManager)
library(tidyverse)
library(magrittr)
library(DESeq2)
library(WGCNA)
library(dplyr)

devtools::install_github("kevinblighe/CorLevelPlot")
library(CorLevelPlot)

# load in DESeq2 object
dds_time <- readRDS("/work/clh162/Data+/Sophie/2025-2026-team-code-sac165/Data+_team/Sophie/dds_time.rds")

# load in column data
col_data <- readRDS("/work/clh162/Data+/Sophie/2025-2026-team-code-sac165/Data+_team/Sophie/col_data.rds")

# load in environmental data
env_data <- readRDS("/work/clh162/Data+/Sophie/2025-2026-team-code-sac165/Data+_team/Sophie/env_data.rds")
env_mat <- env_data[, c("Temp_7d_Avg", "Salinity_7d_Avg", "DO_7d_Avg", "pH_7d_Avg")]
rownames(env_mat) <- rownames(col_data)

# load in gene2cat table
gene2cat <- readRDS("/work/clh162/Data+/Sophie/2025-2026-team-code-sac165/Data+_team/Sophie/gene2cat.rds")

# load in goseq results from Month * Location design
results_time <- readRDS("/work/clh162/Data+/Sophie/2025-2026-team-code-sac165/Data+_team/Sophie/results_time.rds")
goseq_results_time <- readRDS("/work/clh162/Data+/Sophie/2025-2026-team-code-sac165/Data+_team/Sophie/goseq_results_time.rds")
```

```{r}
# transform dds into varianced-stabilized dataset (vsd)
vsd <- vst(dds_time, blind = FALSE)

# extract normalized counts to a matrix and transpose for WGCNA
normalized_counts <- assay(vsd) %>%
  t() # transpose function

# filter to focus on top most variable genes
gene_var <- apply(normalized_counts, 2, var)
top_genes <- names(sort(gene_var, decreasing = TRUE))[1:8000]
normalized_counts <- normalized_counts[, top_genes]
```

# Pick Soft-Threshold
```{r}
# power vector
powers <- c(1:10, seq(12, 20, by = 2))

# run pickSoftThreshold with normalized counts data
sft <- pickSoftThreshold(
  normalized_counts,
  powerVector = powers
)
sft
```

```{r}
# convert sft to dataframe and add new column for model fit 
sft_df <- data.frame(sft$fitIndices) %>%
  dplyr::mutate(model_fit = -sign(slope) * SFT.R.sq) # scale-free networks should have negative slopes; good model fit is positive (model fit = signed R^2)

# plot model fit vs soft-threshold to decide which soft-threshold power is best
ggplot(sft_df, aes(Power, y = model_fit, label = Power)) +
  # plot points
  geom_point() +
  # make power labels slightly above the points
  geom_text(nudge_y = 0.1) +
  # plot line corresponding to recommended wgcna R^2 value cutoff
  geom_hline(yintercept = 0.80, col = "red") +
  # add labels for x and y axis
  xlab("Soft Threshold (power)") +
  ylab("Scale Free Topology Model Fit, signed R^2") + 
  ggtitle("Scale Independence")

# pick power value that is >= 0.80 or if there are multiple that are >= 0.80, pick the one at an inflection point --> I am choosing 5 as the soft-threshold power because it is at the curve of the plot and above 0.80
```

# Run WGCNA
```{r}
# create blockwise network object
bwnet <- blockwiseModules(
  normalized_counts,
  maxBlockSize = 5000, # size chunks of genes the calculations are run in 
  TOMType = "signed", # topological overlap matrix
  power = 5, # use power from scale independence plot
  corFnc = "WGCNA::cor",
  numericLabels = TRUE,
  minModuleSize = 100,
  mergeCutHeight = 0.4
)

# save result object to RDS file
saveRDS(bwnet, file = "/work/clh162/Data+/Sophie/2025-2026-team-code-sac165/Data+_team/Sophie/wgcna_results.rds")
```

# Load WGCNA blockwise object
```{r}
bwnet <- readRDS("wgcna_results.rds")
```

# Module Eigengenes
```{r}
module_eigengenes <- bwnet$MEs

# get number of genes in each module
table(bwnet$colors)
```

```{r}
# plot dendrograms (there are multiple blocks so have to plot them separately in a loop) 
for (i in 1:length(bwnet$dendrograms)) {
 plotDendroAndColors(
  bwnet$dendrograms[[i]],
  cbind(bwnet$unmergedColors[bwnet$blockGenes[[i]]], bwnet$colors[bwnet$blockGenes[[i]]]),
  cbind("unmerged", "merged"),
  dendroLabels = FALSE,
  addGuide = TRUE,
  hang = 0.03,
  guideHang = 0.05,
  main = paste("Block", i)
  ) 
}
```
# Find Correlations with Environmental Variables

I split the heatmaps into four different plots to make the correlations and module eigengenes easier to identify.
```{r}
num_samples <- nrow(normalized_counts)
num_genes <- ncol(normalized_counts)

# find pearson correlation coefficients and p-values
module_env_correlation <- cor(module_eigengenes, env_mat, use = 'p')
module_env_correlation_pvals <- corPvalueStudent(module_env_correlation, num_samples)

# create a trait matrix for environmental variables, month, and location
trait_data <- env_mat

# create binary variables for month 
trait_data$June <- as.numeric(col_data$Month == "June")
trait_data$July <- as.numeric(col_data$Month == "July")
trait_data$August <- as.numeric(col_data$Month == "August")

# create binary variables for location
trait_data$CMT <- as.numeric(col_data$Location == "CMT")
trait_data$DAF <- as.numeric(col_data$Location == "DAF")

# visualize module-trait correlations as heatmap
heatmap_data <- merge(module_eigengenes, trait_data, by = "row.names")
heatmap_data <- heatmap_data %>%
  column_to_rownames(var = "Row.names")

# heatmap
CorLevelPlot(heatmap_data,
             x = names(heatmap_data)[12:20],
             y = names(heatmap_data)[1:11],
             col = c("blue1", "skyblue", "white", "pink", "red")
             )
```
Module 1
```{r}
# create data frame of gene ids and the modules they are in
module_gene_mapping <- as.data.frame(bwnet$colors)

# filter for module 1

# get names of module 1 genes (generic gene ids)
module1_genes <- module_gene_mapping %>%
  filter(`bwnet$colors` == 1) %>%
  rownames()

# filter goseq results for the module genes
module1_gene_symbol <- results_time %>%
  filter(gene_id %in% module1_genes) %>%
  pull(gene_symbol) %>%
  unique()

module1_goseq <- goseq_results_time %>%
  left_join(
    gene2cat %>%
      filter(ensembl_gene_id %in% module1_gene_symbol),
    by = c("category" = "go_id")
  ) %>%
  filter(!is.na(ensembl_gene_id))

module1_goseq <- module1_goseq %>% 
  group_by(category, over_represented_pvalue, under_represented_pvalue, numDEInCat, numInCat, term, ontology, padj) %>%
  summarise(
    n_module_genes = n(),
    module1_genes = paste(sort(unique(ensembl_gene_id)), collapse = ", "),
    .groups = "drop"
  )

module1_goseq

```
Module 5
```{r}
# create data frame of gene ids and the modules they are in
module_gene_mapping <- as.data.frame(bwnet$colors)

# filter for module 5

# get names of module 5 genes (generic gene ids)
module5_genes <- module_gene_mapping %>%
  filter(`bwnet$colors` == 5) %>%
  rownames()

# filter goseq results for the module genes
module5_gene_symbol <- results_time %>%
  filter(gene_id %in% module5_genes) %>%
  pull(gene_symbol) %>%
  unique()

module5_goseq <- goseq_results_time %>%
  left_join(
    gene2cat %>%
      filter(ensembl_gene_id %in% module5_gene_symbol),
    by = c("category" = "go_id")
  ) %>%
  filter(!is.na(ensembl_gene_id))

module5_goseq <- module5_goseq %>% 
  group_by(category, over_represented_pvalue, under_represented_pvalue, numDEInCat, numInCat, term, ontology, padj) %>%
  summarise(
    n_module_genes = n(),
    module5_genes = paste(sort(unique(ensembl_gene_id)), collapse = ", "),
    .groups = "drop"
  )

module5_goseq

```

# Heatmaps by Location

I am subsetting the data by location to see if there are any differences in correlations with environmental variables for each location.
```{r}
env_mat_daf <- env_mat[
  grepl("^DAF", rownames(env_mat)),
]
env_mat_cmt <- env_mat[
  grep("^CMT", rownames(env_mat)),
]

module_eigengenes_daf <- module_eigengenes[
  rownames(env_mat_daf),
]
module_eigengenes_cmt <- module_eigengenes[
  rownames(env_mat_cmt),
]

heatmap_data_daf <- merge(
  module_eigengenes_daf,
  env_mat_daf,
  by = "row.names"
) %>%
  column_to_rownames("Row.names")
heatmap_data_cmt <- merge(
  module_eigengenes_cmt,
  env_mat_cmt,
  by = "row.names"
) %>%
  column_to_rownames("Row.names")
```

```{r}
CorLevelPlot(
  heatmap_data_daf,
  x = names(heatmap_data_daf)[12:15],
  y = names(heatmap_data_daf)[1:11],
  col = c("blue1", "skyblue", "white", "pink", "red"),
  main = "DAF Module-Environment Correlations"
)

CorLevelPlot(
  heatmap_data_cmt,
  x = names(heatmap_data_cmt)[12:15],
  y = names(heatmap_data_cmt)[1:11],
  col = c("blue1", "skyblue", "white", "pink", "red"),
  main = "CMT Module-Environment Correlations"
)
```



# Intramodular Analysis

Using the module membership, you can identify genes with high module membership in interesting modules. 
Using the gene significance, you can identify genes that that have a high significance for each environmental variable of interest.

I am going to identity genes that have a high significance for each environmental variable, then see if those genes have high significance for the modules that have high significance for that environmental variable, and if those same genes have high module membership in those modules. For example, identify a gene that has high significance for temperature, then see if that gene is in module 64 (which we saw had a strong correlation with temperature), and if that gene has high module membership in module 64. 

Module Membership Calculations
```{r}
# calculate module membership, which quantifies how strongly a gene belongs to a specific module
module_membership_measure <- cor(module_eigengenes, normalized_counts, use = 'p')

# calculate p-values for these module membership measures
module_membership_measure_pvals <- corPvalueStudent(module_membership_measure, num_samples)
```

Gene Significance Calculations - Temperature
```{r}
# calculate gene significance and associated p-values for temperature
gene_temp_signif_cor <- cor(normalized_counts, env_mat$Temp_7d_Avg, use = 'p')
gene_temp_signif_cor_pvals <- corPvalueStudent(gene_temp_signif_cor, num_samples)

gene_temp_signif_cor_pvals %>%
  as.data.frame() %>%
  arrange(V1) %>%
  head(25)

gene_temp_signif_cor %>%
  as.data.frame() %>%
  arrange(desc(V1)) %>%
  head(25)
```


Gene Significance Calculations - Salinity
```{r}
# calculate gene significance and associated p-values for salinity
gene_salinity_signif_cor <- cor(normalized_counts, env_mat$Salinity_7d_Avg, use = 'p')
gene_salinity_signif_cor_pvals <- corPvalueStudent(gene_salinity_signif_cor, num_samples)

gene_salinity_signif_cor_pvals %>%
  as.data.frame() %>%
  arrange(V1) %>%
  head(25)
```

Gene Significance Calculations - Dissolved Oxygen
```{r}
# calculate gene significance and associated p-values for dissolved oxygen
gene_do_signif_cor <- cor(normalized_counts, env_mat$DO_7d_Avg, use = 'p')
gene_do_signif_cor_pvals <- corPvalueStudent(gene_do_signif_cor, num_samples)

gene_do_signif_cor_pvals %>%
  as.data.frame() %>%
  arrange(V1) %>%
  head(25)
```

Gene Significance Calculations - pH
```{r}
# calculate gene significance and associated p-values for pH
gene_ph_signif_cor <- cor(normalized_counts, env_mat$pH_7d_Avg, use = 'p')
gene_ph_signif_cor_pvals <- corPvalueStudent(gene_ph_signif_cor, num_samples)

gene_ph_signif_cor_pvals %>%
  as.data.frame() %>%
  arrange(V1) %>%
  head(25)
```


